---
title: "step_anali_spring2026"
author: "Jessica Arend + Anali Kocevar"
date: "2026-02-25"
output:
  word_document: default
  html_document: default
---
# Load, clean, combine data [hidden in output]
```{r, echo = FALSE, include = FALSE}
# clear workspace
rm(list=ls())

# read in packages
pacman::p_load(stats, stringr, dplyr, tidyr, ggplot2, lme4, emmeans, lmerTest, Matrix, pbkrtest, psych, car, performance, parameters, data.table, tibble, Hmisc)
```

## TOPX
```{r}
# load in data
dat_topx <- read.csv("data/aggregate_topx_20260414.csv")
colnames(dat_topx)[1] <- "subj_idx"

# set the levels for variable "visit"
dat_topx$visit <- factor(dat_topx$visit,
                         levels = c("BL1", "BL2", "6M", "7M", "12M"))

# how many ppts total? (n = 201)
length(unique(dat_topx$subj_idx))

# keep only ppts with BL1, BL2, and 6M (n = 105)
proj3_visits <- c("BL1", "BL2", "6M")
dat_topx_proj3 <- dat_topx %>% 
  group_by(subj_idx) %>% 
  filter(all(proj3_visits %in% visit)) %>% 
  ungroup()
length(unique(dat_topx_proj3$subj_idx))
# remove BL2
dat_topx_proj3 <- dat_topx_proj3 %>% 
  filter(visit %in% c("BL1", "6M"))
# set this to dataframe going forward, if using for analyses
dat_topx <- dat_topx_proj3

# keep only ppts with all visits, BL1 through 12M (n = 32)
proj4_visits <- c("BL1", "BL2", "6M", "7M", "12M")
dat_topx_proj4 <- dat_topx %>% 
  group_by(subj_idx) %>% 
  filter(all(proj4_visits %in% visit)) %>% 
  ungroup()
length(unique(dat_topx_proj4$subj_idx))
```

## Demographics
```{r, echo = FALSE, include = FALSE}
### demographics -------
# pull data from REDCap > "JLA Demographics + Dx" report 
# load in and clean data
dat_behav <- read.csv("data/step_behav_dat20251101.csv")
colnames(dat_behav)[1] <- "subj_idx"

# recode race NAs to "other"
dat_behav <- dat_behav %>%
  mutate(race_5split = ifelse(race == 6, 5, race),
         .after = "race")

# create numeric value for group
dat_behav$group_num <- ifelse(dat_behav$group == "Control", 1, 
                     ifelse(dat_behav$group == "Patient", 2, 
                            NA))

# remove bprs, 
dat_behav <- dat_behav %>%
  select(-starts_with("bprs"))
dat_behav <- dat_behav %>%
  select(-starts_with("sans"))
dat_behav <- dat_behav %>%
  select(-starts_with("saps"))
dat_behav <- dat_behav %>%
  select(-starts_with("spq"))

# subset demographics data by topx data
dat_behav <- dat_behav %>%
  filter(subj_idx %in% dat_topx$subj_idx)
```

## BPRS
```{r}



# bprs
dat_behav <- dat_behav %>% 
  mutate(bprs_pos_ws = bprs_pos_ws/4) %>% 
  mutate(bprs_neg_ws = bprs_neg_ws/3) %>% 
  mutate(bprs_disorg_ws = bprs_disorg_ws/6) %>% 
  mutate(bprs_mania_ws = bprs_mania_ws/5) %>%
  mutate(bprs_dep_anx_ws = bprs_dep_anx_ws/5)
```

# DESCRIPTIVES
## Subset data by group/visit
```{r}
# number of patients and controls
table(dat_behav$group)

### create datasets by group ----------
dat_behav_pt <- subset(dat_behav, group == "Patient") 
dat_behav_ctl <- subset(dat_behav, group == "Control")
nPt <- nrow(dat_behav_pt)
nCtl <- nrow(dat_behav_ctl)
```

## Functions
### Count + % table
```{r, echo = FALSE}
# function for count, %, and NA table - 2 split (pt/ctl)
table_ct_pct <- function(var) {
  
  # Calculate distribution and percentages for patients
  count_pt <- table(dat_behav_pt[[var]])
  pct_pt <- round((count_pt/nPt)*100, 0)
  na_pt <- sum(is.na(dat_behav_pt[[var]]))
  
  # Calculate distribution and percentages for controls
  count_ctl <- table(dat_behav_ctl[[var]])
  pct_ctl <- round((count_ctl/nCtl)*100, 0)
  na_ctl <- sum(is.na(dat_behav_ctl[[var]]))
  
  # Create a summary table
  summary <- tibble(
    Group = c(rep("Patient", length(count_pt) + 1), 
            rep("Control", length(count_ctl) + 1)),
    Variable = c(names(count_pt), "NA", 
             names(count_ctl), "NA"),
    Count = c(as.numeric(count_pt), na_pt,
            as.numeric(count_ctl), na_ctl),
    Percentage = c(as.numeric(pct_pt), NA,
                 as.numeric(pct_ctl), NA))
  summary
}
```

### M + SD table
```{r, echo = FALSE}
# function for mean, sd, and NA table - 2 split (pt/ctl)
table_m_sd <- function(var) {
  
  # Calculate mean and standard deviation for patients
  mean_pt <- mean(dat_behav_pt[[var]], na.rm = TRUE)
  sd_pt <- sd(dat_behav_pt[[var]], na.rm = TRUE)
  
  # Calculate mean and standard deviation for controls
  mean_ctl <- mean(dat_behav_ctl[[var]], na.rm = TRUE)
  sd_ctl <- sd(dat_behav_ctl[[var]], na.rm = TRUE)
  
  # Count missing values for patients and controls
  na_pt <- sum(is.na(dat_behav_pt[[var]]))
  na_ctl <- sum(is.na(dat_behav_ctl[[var]]))
  
  # Create a summary table
  summary <- tibble(
    Group = c("Patient", "Control"),
    Mean = c(round(mean_pt, 2), round(mean_ctl, 2)),
    SD = c(round(sd_pt, 2), round(sd_ctl, 2)),
    Missing_Values = c(na_pt, na_ctl))  # Add missing values count
  summary %>% 
    mutate(across(where(is.numeric), as.character))
}
```

### BPRS tables
```{r, echo = FALSE}
# function for BPRS table with wilson-sponheim factors
table_bprs_ws <- function(data) {
  data %>%
    group_by(visit) %>%
    summarise(
      bprs_pos_mean     = round(mean(bprs_pos_ws, na.rm = TRUE), 2),
      bprs_neg_mean     = round(mean(bprs_neg_ws, na.rm = TRUE), 2),
      bprs_disorg_mean  = round(mean(bprs_disorg_ws, na.rm = TRUE), 2),
      bprs_mania_mean   = round(mean(bprs_mania_ws, na.rm = TRUE), 2),
      bprs_dep_anx_mean = round(mean(bprs_dep_anx_ws, na.rm = TRUE), 2),
      
      bprs_pos_sd      = round(sd(bprs_pos_ws, na.rm = TRUE), 2),
      bprs_neg_sd      = round(sd(bprs_neg_ws, na.rm = TRUE), 2),
      bprs_disorg_sd   = round(sd(bprs_disorg_ws, na.rm = TRUE), 2),
      bprs_mania_sd    = round(sd(bprs_mania_ws, na.rm = TRUE), 2),
      bprs_dep_anx_sd  = round(sd(bprs_dep_anx_ws, na.rm = TRUE), 2),
      .groups = "drop"
    ) %>%
    rowwise() %>%
    do({
      tibble(
        visit = .$visit,
        Variable = c("BPRS Positive WS", "BPRS Negative WS", 
                     "BPRS Disorganized WS", "BPRS Mania WS", 
                     "BPRS Depression WS"),
        Mean = c(.$bprs_pos_mean, .$bprs_neg_mean, .$bprs_disorg_mean, 
                 .$bprs_mania_mean, .$bprs_dep_anx_mean),
        SD = c(.$bprs_pos_sd, .$bprs_neg_sd, .$bprs_disorg_sd, 
               .$bprs_mania_sd, .$bprs_dep_anx_sd)
      )
    }) %>%
    ungroup()
}
```

## Demographics
### Gender
```{r}
#gender: 1=man, 2=woman, 3=non-binary/other
table_ct_pct("gender")
```

### Age
```{r}
table_m_sd("age")
```

### Race
```{r}
# race: 0=white, 1=black/african am, 2=indigenous am/alaska native, 3=asian/asian am, 4=native hawaiian/pacific islander, 5=multiracial/other
table_ct_pct("race_5split")
```

### Hispanicity
```{r}
# manually changed SP1030 = yes, SP1097 = no, SP1099 = no
table_ct_pct("hispanic")
```

### Personal Education (Highest)
```{r}
table_m_sd("subj_education")
```

### Caregiver Education (Highest)
```{r}
table_m_sd("hi_caregiver_education")
```

## Clinical 
### Diagnosis
```{r}
# dx: 0=control, 1=Sz/Szform, 2=SzAff, 3=BP, 4=MDD, 5=NOS/other
table_ct_pct("dx_5split")
```

### BPRS - Pt
```{r}
### table of patient BPRS (Wilson-Sponheim factors)
table_bprs_ws(dat_behav_pt)
```

### BPRS - Ctl
```{r}
### table of control BPRS (Wilson-Sponheim factors)
table_bprs_ws(dat_behav_ctl)
```

### BPRS totals
```{r}
# patient bprs total
bprs_total_table_pt <- dat_bprs_pt %>% 
  group_by(visit) %>% 
  dplyr::summarize(Mean = mean(bprs_total, na.rm = TRUE),
            SD = sd(bprs_total, na.rm = TRUE))
print(bprs_total_table_pt)


bprs_total_table_ctl <- dat_bprs_ctl %>% 
  group_by(visit) %>% 
  dplyr::summarize(Mean = mean(bprs_total, na.rm = TRUE),
            SD = sd(bprs_total, na.rm = TRUE))
print(bprs_total_table_ctl)
```

# GROUP DIFF
## Demographics
### Gender
```{r}
# gender - using fisher because samples are too small for chi-squared
fisher.test(dat_behav$gender, dat_behav$group)
```

### Age
```{r}
# age
# welch's two-sample t-test
t.test(age ~ group_num, data = dat_behav)

# test variance/normality to see if t-tests are appropriate
leveneTest(dat_behav$age, dat_behav$group_num, data = dat_behav) # tests equal variance between groups
shapiro.test(dat_behav$age) # tests normality

# wilcox aka mann-whitney U test
wilcox.test(age ~ group_num, data = dat_behav)
```

### Race
```{r}
# race - using fisher because samples are too small for chi-squared
fisher.test(dat_behav$race, dat_behav$group)
```

### Ethnicity
```{r}
# ethnicity
chisq.test(dat_behav$hispanic, dat_behav$group)
```

### Pt Education
```{r}
# participant education
# welch's two-sample t-test
t.test(subj_education ~ group_num, data = dat_behav)

# test variance/normality to see if t-tests are appropriate
leveneTest(dat_behav$subj_education, dat_behav$group_num, data = dat_behav)
shapiro.test(dat_behav$subj_education) # tests normality

wilcox.test(subj_education ~ group_num, data = dat_behav)
```

### Caregiver Education
```{r}
# caregiver education
# welch's two-sample t-test
t.test(hi_caregiver_education ~ group_num, data = dat_behav)

# test variance/normality to see if t-tests are appropriate
leveneTest(dat_behav$hi_caregiver_education, dat_behav$group_num, data = dat_behav)
shapiro.test(dat_behav$hi_caregiver_education) # tests normality

wilcox.test(hi_caregiver_education ~ group_num, data = dat_behav)
```

###IQ
```{r}
# # IQ
# # welch's two-sample t-test
# t.test(iq ~ group_num, data = dat_behav)
# 
# # test variance/normality to see if t-tests are appropriate
# leveneTest(dat_behav$iq, dat_behav$group_num, data = dat_behav)
# wilcox.test(iq ~ group_num, data = dat_behav)
```

## BPRS
### Total
```{r}
# BPRS - total
# welch's two-sample t-test
# t.test(bprs_total ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_total, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_total ~ group_num, data = dat_bprs)

wilcox.test(bprs_total ~ group_num, data = dat_bprs_BL)
wilcox.test(bprs_total ~ group_num, data = dat_bprs_6M)
```

### Positive
```{r}
# BPRS - positive
# welch's two-sample t-test
# t.test(bprs_pos_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_pos_ws, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_pos_ws ~ group_num, data = dat_bprs)

wilcox.test(bprs_pos_ws ~ group_num, data = dat_bprs_BL)
wilcox.test(bprs_pos_ws ~ group_num, data = dat_bprs_6M)
```

### Negative
```{r}
# BPRS - negative
# welch's two-sample t-test
#t.test(bprs_neg_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
#leveneTest(dat_bprs$bprs_neg_ws, dat_bprs$group_num, data = dat_bprs)

wilcox.test(bprs_neg_ws ~ group_num, data = dat_bprs_BL)
wilcox.test(bprs_neg_ws ~ group_num, data = dat_bprs_6M)
```

### Disorg
```{r}
# BPRS - disorg
# welch's two-sample t-test
# t.test(bprs_disorg_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_disorg_ws, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_disorg_ws ~ group_num, data = dat_bprs) 

wilcox.test(bprs_disorg_ws ~ group_num, data = dat_bprs_BL)
wilcox.test(bprs_disorg_ws ~ group_num, data = dat_bprs_6M)
```

### Mania
```{r}
# BPRS - mania
# welch's two-sample t-test
# t.test(bprs_mania_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_mania_ws, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_mania_ws ~ group_num, data = dat_bprs) 

wilcox.test(bprs_mania_ws ~ group_num, data = dat_bprs_BL)
wilcox.test(bprs_mania_ws ~ group_num, data = dat_bprs_6M)
```

### Depr/Anx
```{r}
# BPRS - depr/anx
# welch's two-sample t-test
# t.test(bprs_dep_anx_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_dep_anx_ws, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_dep_anx_ws ~ group_num, data = dat_bprs) 

wilcox.test(bprs_dep_anx_ws ~ group_num, data = dat_bprs_BL)
wilcox.test(bprs_dep_anx_ws ~ group_num, data = dat_bprs_6M)
```

# LONGITUDINAL
## TOPX
### Error
```{r}
# run the LMER, report as an ANOVA
mod_err <- lmer(error ~ group * trial_type * visit + (1|subj), data = dat_topx)
anova_err <- anova(mod_err, type = 3)
anova_err
# run post-hoc tests on LMER main and significant effects
print(emmeans(mod_err, list(pairwise ~ group), adjust = "mvt"))
print(emmeans(mod_err, list(pairwise ~ trial_type), adjust = "mvt"))
print(emmeans(mod_err, list(pairwise ~ group|trial_type), adjust = "mvt"))

### plot error----------
plot_err <- ggplot(dat_topx, 
                        aes(x= trial_type, y=error, group = interaction(group,visit))) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group, linetype = visit), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
  scale_linetype_manual(values = c("solid", "dotted", "dashed"), 
                        labels = c("Visit 1", "Visit 2", "6-Month")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err
```

### Error by Visit
```{r}
dat_topx_bl1 <- filter(dat_topx, visit == "BL1")
dat_topx_bl2 <- filter(dat_topx, visit == "BL2")
dat_topx_6m <- filter(dat_topx, visit == "6M")

### plot BL1 error----------
plot_err_bl1 <- ggplot(dat_topx_bl1, 
                        aes(x= trial_type, y=error, group = group)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BL1 Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP        ")) + 
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_bl1

### plot BL2 error----------
plot_err_bl2 <- ggplot(dat_topx_bl2, 
                        aes(x= trial_type, y=error, group = group)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BL2 Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP        ")) + 
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_bl2

### plot 6M error----------
plot_err_6m <- ggplot(dat_topx_6m, 
                        aes(x= trial_type, y=error, group = group)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "6M Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP        ")) + 
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_6m
```

### Error Aff/NonAff
```{r}
dat_topx <- dat_topx %>%
  left_join(
    dat_behav %>% select(subj, dx_affect),
    by = "subj"
  )

dat_topx <- dat_topx %>%
  mutate(dx_affect = ifelse(group == "Control" & is.na(dx_affect),
                            1,
                            dx_affect))

dat_topx$dx_affect <- factor(dat_topx$dx_affect,
                             levels = c(0, 1, 2),
                             labels = c("Ctl", "Non-Aff", "Aff"))


# run the LMER, report as an ANOVA
mod_err_aff <- lmer(error ~ dx_affect * trial_type * visit + (1|subj), data = dat_topx)
anova_err_aff <- anova(mod_err_aff, type = 3)
anova_err_aff
# run post-hoc tests on LMER main and significant effects
print(emmeans(mod_err_aff, list(pairwise ~ dx_affect), adjust = "mvt"))
print(emmeans(mod_err_aff, list(pairwise ~ trial_type), adjust = "mvt"))
print(emmeans(mod_err_aff, list(pairwise ~ dx_affect|trial_type), adjust = "mvt"))

### plot error----------
plot_err_aff <- ggplot(dat_topx, 
                        aes(x= trial_type, y=error, group = interaction(dx_affect,visit))) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = dx_affect, linetype = visit), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
                     labels=c("Ctl", "Non-Aff", "Aff")) + 
  scale_linetype_manual(values = c("solid", "dotted", "dashed"), 
                        labels = c("Visit 1", "Visit 2", "6-Month")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_aff
```

### Error Aff/Nonaff by Visit
```{r}
dat_topx_bl1 <- filter(dat_topx, visit == "BL1")
dat_topx_bl2 <- filter(dat_topx, visit == "BL2")
dat_topx_6m <- filter(dat_topx, visit == "6M")

### plot BL1 error----------
plot_err_bl1_aff <- ggplot(dat_topx_bl1, 
                        aes(x= trial_type, y=error, group = dx_affect)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = dx_affect), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BL1 Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
                     labels=c("Ctl", "Non-Aff", "Aff")) +
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_bl1_aff

### plot BL2 error----------
plot_err_bl2_aff <- ggplot(dat_topx_bl2, 
                        aes(x= trial_type, y=error, group = dx_affect)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = dx_affect), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BL2 Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
                     labels=c("Ctl", "Non-Aff", "Aff")) +
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_bl2_aff

### plot 6M error----------
plot_err_6m_aff <- ggplot(dat_topx_6m, 
                        aes(x= trial_type, y=error, group = dx_affect)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = dx_affect), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "6M Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
                     labels=c("Ctl", "Non-Aff", "Aff")) +
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_6m_aff
```

### Reaction Time
```{r}
# run the LMER, report as an ANOVA
mod_rt <- lmer(rt_mdn ~ group * trial_type * visit + (1|subj), data = dat_topx)
anova_rt <- anova(mod_rt, type = 3)
anova_rt

# run post-hoc tests on LMER main and significant effects
print(emmeans(mod_rt, list(pairwise ~ group), adjust = "mvt"))
print(emmeans(mod_rt, list(pairwise ~ trial_type), adjust = "mvt"))
print(emmeans(mod_rt, list(pairwise ~ visit), adjust = "mvt"))
print(emmeans(mod_rt, list(pairwise ~ group|trial_type), adjust = "mvt"))

### plot reaction time ----------
plot_rt <- ggplot(dat_topx, aes(x= trial_type, y=rt_mdn, group = interaction(group,visit))) +
  stat_summary(fun="mean", geom="line", aes(color = group, linetype = visit), linewidth = 1.5, position = position_dodge(0.2)) +
  stat_summary(fun="mean", geom="point", aes(color = group), size = 3, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "Median Reaction Times (ms)", 
       y = "Median Reaction Times (ms)", 
       x = "Trial Type", 
       color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(250, 600)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
  scale_linetype_manual(values = c("solid", "dotted", "dashed"), 
                        labels = c("Visit 1", "Visit 2", "6-Month")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_rt
```

## BPRS
### Total
```{r}
# run the LMER, report as an ANOVA
mod_bprs <- lmer(bprs_total ~ group * visit + (1|subj), data = dat_bprs)
anova_bprs <- anova(mod_bprs, type = 3)
anova_bprs

# run post-hoc tests on significant effects
print(emmeans(mod_bprs, list(pairwise ~ group), adjust = "mvt"))
```

### Positive
```{r}
# run the LMER, report as an ANOVA
mod_pos <- lmer(bprs_pos_ws ~ group * visit + (1|subj), data = dat_bprs)
anova_pos <- anova(mod_pos, type = 3)
anova_pos

# run post-hoc tests on significant effects
print(emmeans(mod_pos, list(pairwise ~ group), adjust = "mvt"))
```

### Negative
```{r}
# run the LMER, report as an ANOVA
mod_neg <- lmer(bprs_neg_ws ~ group * visit + (1|subj), data = dat_bprs)
anova_neg <- anova(mod_neg, type = 3)
anova_neg

# run post-hoc tests on significant effects
print(emmeans(mod_neg, list(pairwise ~ group), adjust = "mvt"))
print(emmeans(mod_neg, list(pairwise ~ visit), adjust = "mvt"))
print(emmeans(mod_neg, list(pairwise ~ group|visit), adjust = "mvt"))

### plot BPRS negative ----------
plot_neg <- ggplot(dat_bprs, aes(x= visit, y=bprs_neg_ws)) +
  stat_summary(fun="mean", geom="line", aes(color = group), linewidth = 1.5, position = position_dodge(0.2)) +
  stat_summary(fun="mean", geom="point", aes(color = group), size = 3, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BPRS Negative", 
       y = "BPRS Negative", 
       x = "Visit",
       color = "Group") +
  coord_cartesian(ylim = c(0, 10)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_neg
```

### Disorg
```{r}
# run the LMER, report as an ANOVA
mod_disorg <- lmer(bprs_disorg_ws ~ group * visit + (1|subj), data = dat_bprs)
anova_disorg <- anova(mod_disorg, type = 3)
anova_disorg

# run post-hoc tests on significant effects
print(emmeans(mod_disorg, list(pairwise ~ group), adjust = "mvt"))
```

### Plot All
```{r}
# convert BPRS from wide to long format
dat_bprs_long <- dat_bprs %>% 
  select(c(subj, visit, group, group_num, bprs_pos_ws, bprs_neg_ws,
           bprs_disorg_ws, bprs_mania_ws, bprs_dep_anx_ws)) %>% 
  pivot_longer(cols = c(bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws, 
                        bprs_mania_ws, bprs_dep_anx_ws), 
               names_to = "bprs_symptom", 
               values_to = "bprs_score") %>% 
  mutate(bprs_symptom = dplyr::recode(bprs_symptom,
         "bprs_pos_ws" = "Pos",
         "bprs_neg_ws" = "Neg",
         "bprs_disorg_ws" = "Disorg",
         "bprs_mania_ws" = "Mania",
         "bprs_dep_anx_ws" = "Dep-Anx"))

# Lineplot of BPRS all symptoms
plot_bprs <- ggplot(dat_bprs_long, 
                        aes(x= bprs_symptom, y=bprs_score, group = interaction(group,visit))) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group, linetype = visit), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BPRS symptoms", 
       y = "BPRS symptom rating", 
       x = "BPRS subfactor", 
       color = "Group", linetype = "Visit") +
  #coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
  scale_linetype_manual(values = c("solid", "dotted"), 
                        labels = c("Baseline", "6-Month")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_bprs

# Barplot of BPRS all symptom
plot_bprs_bar <- ggplot(
  dat_bprs_long,
  aes(x = bprs_symptom, y = bprs_score, fill = group)
) +
  stat_summary(
    fun = "mean",
    geom = "bar",
    position = position_dodge(width = 0.9),
    color = "black"
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    position = position_dodge(width = 0.9),
    width = 0.3
  ) +
  facet_wrap(~ visit) +
  labs(
    title = "BPRS Symptoms",
    y = "BPRS Symptom Rating",
    x = "BPRS Subfactor",
    fill = "Group"
  ) +
  scale_fill_manual(
    values = c("#2C737F", "#AA8829"),
    labels = c("Ctl", "EP")
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 20),
    legend.text = element_text(face = "bold", size = 14),
    legend.title = element_text(face = "bold", size = 16),
    plot.title = element_text(face = "bold", size = 22)
  )
plot_bprs_bar
```

# CORRELATIONS
## Test Normality
```{r}
# Test normal distribution of variables
# if p < .05, variable is non-normal/skewed 
shapiro.test(dat_all$error_AX)
shapiro.test(dat_all$error_AY)
shapiro.test(dat_all$error_BX)
shapiro.test(dat_all$error_BY)
shapiro.test(dat_all$bprs_pos_ws)
shapiro.test(dat_all$bprs_neg_ws)
shapiro.test(dat_all$bprs_disorg_ws)
```

## Create Dataframe
```{r}
# create new bprs dataframe with only necessary variables
dat_bprs_abbrev <- dat_bprs %>% 
  select(subj, visit, bprs_total, bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws, bprs_mania_ws, bprs_dep_anx_ws)

# create new topx dataframe, eliminating BL2
dat_topx_abbrev <- dat_topx %>% 
  filter(visit == "BL1" | visit == "6M") 
dat_topx_abbrev <- dat_topx_abbrev %>% 
  select(subj, visit, group, d_context, d_expect, trial_type, rt_mdn, error)
dat_topx_abbrev$visit <- as.character(dplyr::recode(dat_topx_abbrev$visit, "BL1" = "BL"))
dat_topx_abbrev <- pivot_wider(dat_topx_abbrev, 
                               id_cols = c(subj, visit, group, 
                                           d_context, d_expect), 
                               values_from = c(rt_mdn, error), 
                               names_from = trial_type)

# combine topx & bprs data
dat_all <- dat_topx_abbrev %>%
  left_join(dat_bprs_abbrev %>% 
              select(subj, visit, bprs_total, bprs_pos_ws, 
                     bprs_neg_ws, bprs_disorg_ws, bprs_mania_ws, 
                     bprs_dep_anx_ws),
    by = c("subj", "visit"))

# create dataframes for each visit
dat_all_bl <- dat_all %>% 
  filter(visit == "BL")
dat_all_6m <- dat_all %>% 
  filter(visit == "6M")
```

## TOPX + BPRS
```{r}
# create dataframe for a correlation matrix of TOPX error rates and BPRS symptoms
dat_cor1 <- dat_all %>% 
  select(error_AX, error_AY, error_BX, bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws)

# compute and print correlations + p-values
cor_results <- rcorr(as.matrix(dat_cor1), type = "pearson")
(cor_matrix <- cor_results$r)
(p_matrix   <- cor_results$P)
```
