---
title: "STEP Longitudinal Analyses of Clinical/Cognitive Measures"
author: "Jessica Arend"
date: "2026-04-14"
output:
  html_document: default
  word_document: default
---
# Load, clean, combine data [hidden in output]
```{r, echo = FALSE, include = FALSE}
# clear workspace
rm(list=ls())

# read in packages
pacman::p_load(stats, stringr, dplyr, tidyr, ggplot2, ggpattern, lme4, emmeans, lmerTest, Matrix, pbkrtest, psych, car, performance, parameters, data.table, tibble, Hmisc, corrplot)
```

## Demographics
```{r, echo = FALSE, include = FALSE}
### demographics -------
# pull data from REDCap > "JLA Demographics + Dx" report 
# load in and clean data
dat_demo <- read.csv("data/step_demographics_clean_20260414.csv")
colnames(dat_demo)[1] <- "subj"

# recode race NAs to "other"
dat_demo <- dat_demo %>%
  mutate(race_5split = ifelse(race == 6, 5, race),
         .after = "race")

dat_demo <- dat_demo %>% 
  mutate(visit = "BL1", .after = subj)
```

## TOPX
```{r}
# load in data
dat_topx <- read.csv("data/aggregate_topx_20260414.csv")
# remove duplicate 6M visit for SP2150
dat_topx <- filter(dat_topx, !(subj == "SP2150" & visit == "6M" & date == "20231130"))

# how many ppts total? (n = 201)
length(unique(dat_topx$subj))

dat_topx <- dat_topx %>%
  rename(group = subj_type) %>%
  mutate(
    group_num = case_when(
      group == "Control" ~ 0,
      group == "Patient" ~ 1))

# create project 3 dataset (n = 105)
proj3_visits <- c("BL1", "BL2", "6M")
dat_topx_proj3 <- dat_topx %>% 
  group_by(subj) %>% 
  # remove 7M, 12M
  filter(visit %in% c("BL1", "BL2", "6M"))  %>% 
  # only keep ppl with all three visits: BL1, BL2, 6M
  filter(all(proj3_visits %in% visit)) %>% 
  ungroup() 
length(unique(dat_topx_proj3$subj))
table(dat_topx_proj3$visit)
# set factor to order visits
dat_topx_proj3$visit <- factor(dat_topx_proj3$visit,
                         levels = c("BL1", "BL2", "6M"))

# create project 3 dataset without BL2
dat_topx_proj3_no_bl2 <- dat_topx_proj3 %>% 
  filter(visit %in% c("BL1", "6M"))
dat_topx_proj3_no_bl2$visit <- factor(dat_topx_proj3_no_bl2$visit,
                         levels = c("BL1","6M"))
# set this to dataframe going forward, if using for analyses
# dat_topx <- dat_topx_proj3

# keep only ppts with all visits, BL1 through 12M (n = 32)
proj4_visits <- c("BL1", "BL2", "6M", "7M", "12M")
dat_topx_proj4 <- dat_topx %>% 
  group_by(subj) %>% 
  filter(all(proj4_visits %in% visit)) %>% 
  ungroup()
length(unique(dat_topx_proj4$subj))
# set the levels for variable "visit"
dat_topx_proj4$visit <- factor(dat_topx_proj4$visit,
                         levels = c("BL1", "BL2", "6M", "7M", "12M"))

# subset demographics data by topx data
dat_demo <- dat_demo %>%
  filter(subj %in% dat_topx$subj)
```

## Test My Brain DNT
```{r}
dat_tmb <- read.csv("../data/nontask_data/other_data/step_tmb_20260928.csv")

# subset tmb data by topx data
dat_tmb <- dat_tmb %>%
  filter(subj %in% dat_topx$subj) %>%
  mutate(
    group_num = case_when(
      group == "Control" ~ 0,
      group == "Patient" ~ 1))
```

## Clinical Symptoms
### BPRS
```{r}
# read in bprs data
dat_bprs <- read.csv("data/step_bprs_clean_20260417.csv")

#rename variables 
colnames(dat_bprs)[1] <- "subj"
dat_bprs <- dat_bprs %>%
  rename(bprs_pos_ws = bprs_pos_ws_avg,
         bprs_neg_ws = bprs_neg_ws_avg,
         bprs_disorg_ws = bprs_disorg_ws_avg,
         bprs_mania_ws = bprs_mania_ws_avg,
         bprs_dep_anx_ws = bprs_dep_anx_ws_avg)

# add variable for group
dat_bprs$group = "group"
dat_bprs <- dat_bprs %>% 
  mutate(
    group = case_when(
      str_sub(subj, 1, 3) == 'SP1' ~ 'Control',
      str_sub(subj, 1, 3) == 'SP2' ~ 'Patient',
      TRUE ~ group)) %>% 
  mutate(
    group_num = case_when(
      group == "Control" ~ 0,
      group == "Patient" ~ 1))

# subset bprs data by topx data
dat_bprs <- dat_bprs %>%
  filter(subj %in% dat_topx$subj)
```

### SANS-SAPS
```{r}
# read in sans-saps data
dat_sanssaps <- read.csv("data/step_sanssaps_clean_20260414.csv")

# rename variables
colnames(dat_sanssaps)[1] <- "subj"

# add variable for group
dat_sanssaps$group = "group"
dat_sanssaps <- dat_sanssaps %>% 
  mutate(
    group = case_when(
      str_sub(subj, 1, 3) == 'SP1' ~ 'Control',
      str_sub(subj, 1, 3) == 'SP2' ~ 'Patient',
      TRUE ~ group)) %>% 
  mutate(
    group_num = case_when(
      group == "Control" ~ 0,
      group == "Patient" ~ 1))

# subset sans-saps by topx data
dat_sanssaps <- dat_sanssaps %>%
  filter(subj %in% dat_topx$subj)
```

### Global functioning
```{r}
# read in global functioning data
dat_gf <- read.csv("data/step_gf_clean_20260417.csv")
colnames(dat_gf)[1] <- "subj"

# add variable for group
dat_gf$group = "group"
dat_gf <- dat_gf %>% 
  mutate(
    group = case_when(
      str_sub(subj, 1, 3) == 'SP1' ~ 'Control',
      str_sub(subj, 1, 3) == 'SP2' ~ 'Patient',
      TRUE ~ group)) %>% 
  mutate(
    group_num = case_when(
      group == "Control" ~ 0,
      group == "Patient" ~ 1))

# subset global functioning data by topx data
dat_gf <- dat_gf %>%
  filter(subj %in% dat_topx$subj)
```

# DESCRIPTIVES
## Subset data by group/visit
```{r}
# number of patients and controls
table(dat_demo$group)

### create datasets by group/timepoint----------
# demographics
dat_demo_pt <- subset(dat_demo, group == "Patient") 
dat_demo_ctl <- subset(dat_demo, group == "Control")
# count ppts
nPt <- nrow(dat_demo_pt)
nCtl <- nrow(dat_demo_ctl)

# bprs
dat_bprs_pt <- subset(dat_bprs, group == "Patient") 
dat_bprs_ctl <- subset(dat_bprs, group == "Control")
dat_bprs_bl <- subset(dat_bprs, visit != "6M") 
dat_bprs_6m <- subset(dat_bprs, visit == "6M")

# count ppts
nPt <- nrow(dat_demo_pt)
nCtl <- nrow(dat_demo_ctl)
```

## Functions
### Count + % table
```{r, echo = FALSE}
# function for count, %, and NA table - 2 split (pt/ctl)
table_ct_pct <- function(var) {
  
  # Calculate distribution and percentages for patients
  count_pt <- table(dat_demo_pt[[var]])
  pct_pt <- round((count_pt/nPt)*100, 0)
  na_pt <- sum(is.na(dat_demo_pt[[var]]))
  
  # Calculate distribution and percentages for controls
  count_ctl <- table(dat_demo_ctl[[var]])
  pct_ctl <- round((count_ctl/nCtl)*100, 0)
  na_ctl <- sum(is.na(dat_demo_ctl[[var]]))
  
  # Create a summary table
  summary <- tibble(
    Group = c(rep("Patient", length(count_pt) + 1), 
            rep("Control", length(count_ctl) + 1)),
    Variable = c(names(count_pt), "NA", 
             names(count_ctl), "NA"),
    Count = c(as.numeric(count_pt), na_pt,
            as.numeric(count_ctl), na_ctl),
    Percentage = c(as.numeric(pct_pt), NA,
                 as.numeric(pct_ctl), NA))
  summary
}
```

### M + SD table
```{r, echo = FALSE}
# function for mean, sd, and NA table - 2 split (pt/ctl)
table_m_sd <- function(var) {
  
  # Calculate mean and standard deviation for patients
  mean_pt <- mean(dat_demo_pt[[var]], na.rm = TRUE)
  sd_pt <- sd(dat_demo_pt[[var]], na.rm = TRUE)
  
  # Calculate mean and standard deviation for controls
  mean_ctl <- mean(dat_demo_ctl[[var]], na.rm = TRUE)
  sd_ctl <- sd(dat_demo_ctl[[var]], na.rm = TRUE)
  
  # Count missing values for patients and controls
  na_pt <- sum(is.na(dat_demo_pt[[var]]))
  na_ctl <- sum(is.na(dat_demo_ctl[[var]]))
  
  # Create a summary table
  summary <- tibble(
    Group = c("Patient", "Control"),
    Mean = c(round(mean_pt, 2), round(mean_ctl, 2)),
    SD = c(round(sd_pt, 2), round(sd_ctl, 2)),
    Missing_Values = c(na_pt, na_ctl))  # Add missing values count
  summary %>% 
    mutate(across(where(is.numeric), as.character))
}
```

### BPRS tables
```{r, echo = FALSE}
# function for BPRS table with wilson-sponheim factors
table_bprs_ws <- function(data) {
  data %>%
    group_by(visit) %>%
    summarise(
      bprs_pos_mean     = round(mean(bprs_pos_ws, na.rm = TRUE), 2),
      bprs_neg_mean     = round(mean(bprs_neg_ws, na.rm = TRUE), 2),
      bprs_disorg_mean  = round(mean(bprs_disorg_ws, na.rm = TRUE), 2),
      bprs_mania_mean   = round(mean(bprs_mania_ws, na.rm = TRUE), 2),
      bprs_dep_anx_mean = round(mean(bprs_dep_anx_ws, na.rm = TRUE), 2),
      
      bprs_pos_sd      = round(sd(bprs_pos_ws, na.rm = TRUE), 2),
      bprs_neg_sd      = round(sd(bprs_neg_ws, na.rm = TRUE), 2),
      bprs_disorg_sd   = round(sd(bprs_disorg_ws, na.rm = TRUE), 2),
      bprs_mania_sd    = round(sd(bprs_mania_ws, na.rm = TRUE), 2),
      bprs_dep_anx_sd  = round(sd(bprs_dep_anx_ws, na.rm = TRUE), 2),
      .groups = "drop"
    ) %>%
    rowwise() %>%
    do({
      tibble(
        visit = .$visit,
        Variable = c("BPRS Positive WS", "BPRS Negative WS", 
                     "BPRS Disorganized WS", "BPRS Mania WS", 
                     "BPRS Depression WS"),
        Mean = c(.$bprs_pos_mean, .$bprs_neg_mean, .$bprs_disorg_mean, 
                 .$bprs_mania_mean, .$bprs_dep_anx_mean),
        SD = c(.$bprs_pos_sd, .$bprs_neg_sd, .$bprs_disorg_sd, 
               .$bprs_mania_sd, .$bprs_dep_anx_sd)
      )
    }) %>%
    ungroup()
}
```

## Demographics
### Gender
```{r}
#gender: 1=man, 2=woman, 3=non-binary/other
table_ct_pct("gender")
```

### Age
```{r}
table_m_sd("age")
range(dat_demo$age)
```

### Race
```{r}
# race: 0=white, 1=black/african am, 2=indigenous am/alaska native, 3=asian/asian am, 4=native hawaiian/pacific islander, 5=multiracial/other
table_ct_pct("race_5split")
```

### Hispanicity
```{r}
# manually changed SP1030 = yes, SP1097 = no, SP1099 = no
table_ct_pct("hispanic")
```

### Personal Education (Highest)
```{r}
table_m_sd("subj_education")
```

### Caregiver Education (Highest)
```{r}
table_m_sd("hi_caregiver_education")
```

## Clinical 
### Diagnosis
```{r}
# dx: 0=control, 1=Sz/Szform, 2=SzAff, 3=BP, 4=MDD, 5=NOS/other
table_ct_pct("dx_5split")

# 0 = Control, 1 = Non-Affective, 2 = Affective
table_ct_pct("dx_affect")
```

### BPRS - Pt
```{r}
### table of patient BPRS (Wilson-Sponheim factors)
table_bprs_ws(dat_bprs_pt)
```

### BPRS - Ctl
```{r}
### table of control BPRS (Wilson-Sponheim factors)
table_bprs_ws(dat_bprs_ctl)
```

### BPRS totals
```{r}
# patient bprs total
bprs_total_table_pt <- dat_bprs_pt %>% 
  group_by(visit) %>% 
  dplyr::summarize(Mean = mean(bprs_total, na.rm = TRUE),
            SD = sd(bprs_total, na.rm = TRUE))
print(bprs_total_table_pt)


bprs_total_table_ctl <- dat_bprs_ctl %>% 
  group_by(visit) %>% 
  dplyr::summarize(Mean = mean(bprs_total, na.rm = TRUE),
            SD = sd(bprs_total, na.rm = TRUE))
print(bprs_total_table_ctl)
```

# GROUP DIFF
## Demographics
### Gender
```{r}
# gender - using fisher because samples are too small for chi-squared
fisher.test(dat_demo$gender, dat_demo$group)
```

### Age
```{r}
# age
# welch's two-sample t-test
t.test(age ~ group_num, data = dat_demo)

# test variance/normality to see if t-tests are appropriate
leveneTest(dat_demo$age, dat_demo$group_num, data = dat_demo) # tests equal variance between groups
shapiro.test(dat_demo$age) # tests normality

# wilcox aka mann-whitney U test
wilcox.test(age ~ group_num, data = dat_demo)
```

### Race
```{r}
# race - using fisher because samples are too small for chi-squared
fisher.test(dat_demo$race, dat_demo$group)
```

### Ethnicity
```{r}
# ethnicity
chisq.test(dat_demo$hispanic, dat_demo$group)
```

### Pt Education
```{r}
# participant education
# welch's two-sample t-test
t.test(subj_education ~ group_num, data = dat_demo)

# test variance/normality to see if t-tests are appropriate
leveneTest(dat_demo$subj_education, dat_demo$group_num, data = dat_demo)
shapiro.test(dat_demo$subj_education) # tests normality

wilcox.test(subj_education ~ group_num, data = dat_demo)
```

### Caregiver Education
```{r}
# caregiver education
# welch's two-sample t-test
t.test(hi_caregiver_education ~ group_num, data = dat_demo)

# test variance/normality to see if t-tests are appropriate
leveneTest(dat_demo$hi_caregiver_education, dat_demo$group_num, data = dat_demo)
shapiro.test(dat_demo$hi_caregiver_education) # tests normality

wilcox.test(hi_caregiver_education ~ group_num, data = dat_demo)
```

###IQ
```{r}
# # IQ
# # welch's two-sample t-test
# t.test(iq ~ group_num, data = dat_demo)
# 
# # test variance/normality to see if t-tests are appropriate
# leveneTest(dat_demo$iq, dat_demo$group_num, data = dat_demo)
# wilcox.test(iq ~ group_num, data = dat_demo)
```

## BPRS
### Total
```{r}
# BPRS - total
# welch's two-sample t-test
# t.test(bprs_total ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_total, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_total ~ group_num, data = dat_bprs)

wilcox.test(bprs_total ~ group_num, data = dat_bprs_bl)
wilcox.test(bprs_total ~ group_num, data = dat_bprs_6m)
```

### Positive
```{r}
# BPRS - positive
# welch's two-sample t-test
# t.test(bprs_pos_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_pos_ws, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_pos_ws ~ group_num, data = dat_bprs)

wilcox.test(bprs_pos_ws ~ group_num, data = dat_bprs_bl)
wilcox.test(bprs_pos_ws ~ group_num, data = dat_bprs_6m)
```

### Negative
```{r}
# BPRS - negative
# welch's two-sample t-test
#t.test(bprs_neg_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
#leveneTest(dat_bprs$bprs_neg_ws, dat_bprs$group_num, data = dat_bprs)

wilcox.test(bprs_neg_ws ~ group_num, data = dat_bprs_bl)
wilcox.test(bprs_neg_ws ~ group_num, data = dat_bprs_6m)
```

### Disorg
```{r}
# BPRS - disorg
# welch's two-sample t-test
# t.test(bprs_disorg_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_disorg_ws, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_disorg_ws ~ group_num, data = dat_bprs) 

wilcox.test(bprs_disorg_ws ~ group_num, data = dat_bprs_bl)
wilcox.test(bprs_disorg_ws ~ group_num, data = dat_bprs_6m)
```

### Mania
```{r}
# BPRS - mania
# welch's two-sample t-test
# t.test(bprs_mania_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_mania_ws, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_mania_ws ~ group_num, data = dat_bprs) 

wilcox.test(bprs_mania_ws ~ group_num, data = dat_bprs_bl)
wilcox.test(bprs_mania_ws ~ group_num, data = dat_bprs_6m)
```

### Depr/Anx
```{r}
# BPRS - depr/anx
# welch's two-sample t-test
# t.test(bprs_dep_anx_ws ~ group_num, data = dat_bprs)

# test variance/normality to see if t-tests are appropriate
# leveneTest(dat_bprs$bprs_dep_anx_ws, dat_bprs$group_num, data = dat_bprs)
# wilcox.test(bprs_dep_anx_ws ~ group_num, data = dat_bprs) 

wilcox.test(bprs_dep_anx_ws ~ group_num, data = dat_bprs_bl)
wilcox.test(bprs_dep_anx_ws ~ group_num, data = dat_bprs_6m)
```

# LONGITUDINAL
## TOPX
### Error
```{r}
# set this to dataframe going forward, if using for analyses
dat_topx <- dat_topx_proj3_no_bl2

# run the LMER, report as an ANOVA
mod_err <- lmer(error ~ group * trial_type * visit + (1|subj), data = dat_topx)
anova_err <- anova(mod_err, type = 3)
anova_err
# run post-hoc tests on LMER main and significant effects
print(emmeans(mod_err, list(pairwise ~ group), adjust = "mvt"))
print(emmeans(mod_err, list(pairwise ~ trial_type), adjust = "mvt"))
print(emmeans(mod_err, list(pairwise ~ group|trial_type), adjust = "mvt"))

### plot error----------
plot_err <- ggplot(dat_topx, 
                        aes(x= trial_type, y=error, group = interaction(group,visit))) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group, linetype = visit), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
  scale_linetype_manual(values = c("solid", "dotted", "dashed"), 
                        labels = c("Visit 1", "Visit 2", "6-Month")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err
```

### Error by Visit
```{r}
dat_topx_bl1 <- filter(dat_topx, visit == "BL1")
dat_topx_bl2 <- filter(dat_topx, visit == "BL2")
dat_topx_6m <- filter(dat_topx, visit == "6M")

### plot BL1 error----------
plot_err_bl1 <- ggplot(dat_topx_bl1, 
                        aes(x= trial_type, y=error, group = group)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BL1 Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_bl1

### plot BL2 error----------
plot_err_bl2 <- ggplot(dat_topx_bl2, 
                        aes(x= trial_type, y=error, group = group)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BL2 Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
 coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_bl2

### plot 6M error----------
plot_err_6m <- ggplot(dat_topx_6m, 
                        aes(x= trial_type, y=error, group = group)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = group), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = group), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.2, size = 1, position = position_dodge(0.2)) + 
  labs(title = "6M Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_6m
```

### Error Aff/NonAff Linegraph combined
```{r}
dat_topx <- dat_topx %>%
  left_join(
    dat_demo %>% select(subj, dx_affect),
    by = "subj"
  )

dat_topx <- dat_topx %>%
  mutate(dx_affect = ifelse(group == "Control" & is.na(dx_affect),
                            1,
                            dx_affect))

dat_topx$dx_affect <- factor(dat_topx$dx_affect,
                             levels = c(0, 1, 2),
                             labels = c("Ctl", "Non-Aff", "Aff"))


# run the LMER, report as an ANOVA
mod_err_aff <- lmer(error ~ dx_affect * trial_type * visit + (1|subj), data = dat_topx)
anova_err_aff <- anova(mod_err_aff, type = 3)
anova_err_aff
# run post-hoc tests on LMER main and significant effects
print(emmeans(mod_err_aff, list(pairwise ~ dx_affect), adjust = "mvt"))
print(emmeans(mod_err_aff, list(pairwise ~ trial_type), adjust = "mvt"))
print(emmeans(mod_err_aff, list(pairwise ~ dx_affect|trial_type), adjust = "mvt"))

### plot error----------
plot_err_aff <- ggplot(dat_topx, 
                        aes(x= trial_type, y=error, group = interaction(dx_affect,visit))) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = dx_affect, linetype = visit), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(0.05, .4)) +
  scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
                     labels=c("Ctl", "Non-Aff", "Aff")) + 
  scale_linetype_manual(values = c("solid", "dotted", "dashed"), 
                        labels = c("Visit 1", "Visit 2", "6-Month")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_aff
```

### Error Aff/Nonaff Barplot
```{r}
plot_err_aff_bar <- ggplot(
  dat_topx,
  aes(
    x = trial_type,
    y = error,
    fill = dx_affect,
    pattern = visit
  )
) +
  stat_summary(
    fun = "mean",
    geom = "bar_pattern",
    position = position_dodge(width = 0.9),
    color = "black",
    pattern_fill = "black",
    pattern_color = "black",
    pattern_density = 0.1,
    pattern_spacing = 0.02
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    position = position_dodge(width = 0.9),
    width = 0.3
  ) +
  coord_cartesian(ylim = c(0.05, .45)) +
  labs(
    title = "TOPX Error Rates",
    y = "TOPX Error Rate",
    x = "Trial Type",
    fill = "Group",
    pattern = "Visit"
  ) +
  scale_fill_manual(
    values = c("Ctl" = "#2C737F", "Non-Aff" = "#AA8829", "Aff" = "#C44E52"),
    labels = c("Ctl", "Non-Aff", "Aff")
  ) +
  scale_pattern_manual(
    values = c("BL1" = "none", "6M" = "stripe"),
    labels = c("Visit 1", "6-Month")
  ) +
  guides(
    fill = guide_legend(override.aes = list(pattern = "none")),
    pattern = guide_legend()
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 16),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 16),
    plot.title = element_text(face = "bold", size = 22)
  )
plot_err_aff_bar

ggsave(paste0("graphs/", format(Sys.Date(), "%Y%m%d"),
              "_topx_err_affect_barplot.png"),
       plot_err_aff_bar, width = 6, height = 4)
```


### Error Aff/Nonaff by Visit
```{r}
dat_topx_bl1 <- filter(dat_topx, visit == "BL1")
dat_topx_bl2 <- filter(dat_topx, visit == "BL2")
dat_topx_6m <- filter(dat_topx, visit == "6M")

### plot BL1 error----------
plot_err_bl1_aff <- ggplot(dat_topx_bl1, 
                        aes(x= trial_type, y=error, group = dx_affect)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = dx_affect), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
               width = 0.4, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BL1 Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .45)) +
  scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
                     labels=c("Ctl", "Non-Aff", "Aff")) +
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_bl1_aff

### plot BL2 error----------
plot_err_bl2_aff <- ggplot(dat_topx_bl2, 
                        aes(x= trial_type, y=error, group = dx_affect)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = dx_affect), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
               width = 0.4, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BL2 Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .45)) +
  scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
                     labels=c("Ctl", "Non-Aff", "Aff")) +
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_bl2_aff

### plot 6M error----------
plot_err_6m_aff <- ggplot(dat_topx_6m, 
                        aes(x= trial_type, y=error, group = dx_affect)) + 
  stat_summary(fun="mean", geom="line", 
               aes(color = dx_affect), linewidth = 1, position = position_dodge(0.2)) + 
  stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
               width = 0.4, size = 1, position = position_dodge(0.2)) + 
  labs(title = "6M Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group") +
  coord_cartesian(ylim = c(0.05, .45)) +
  scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
                     labels=c("Ctl", "Non-Aff", "Aff")) +
theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_err_6m_aff
```

### Reaction Time
```{r}
# run the LMER, report as an ANOVA
mod_rt <- lmer(rt_mdn ~ group * trial_type * visit + (1|subj), data = dat_topx)
anova_rt <- anova(mod_rt, type = 3)
anova_rt

# run post-hoc tests on LMER main and significant effects
print(emmeans(mod_rt, list(pairwise ~ group), adjust = "mvt"))
print(emmeans(mod_rt, list(pairwise ~ trial_type), adjust = "mvt"))
print(emmeans(mod_rt, list(pairwise ~ visit), adjust = "mvt"))
print(emmeans(mod_rt, list(pairwise ~ group|trial_type), adjust = "mvt"))

### plot reaction time ----------
plot_rt <- ggplot(dat_topx, aes(x= trial_type, y=rt_mdn, group = interaction(group,visit))) +
  stat_summary(fun="mean", geom="line", aes(color = group, linetype = visit), linewidth = 1.5, position = position_dodge(0.2)) +
  stat_summary(fun="mean", geom="point", aes(color = group), size = 3, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "Median Reaction Times (ms)", 
       y = "Median Reaction Times (ms)", 
       x = "Trial Type", 
       color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(250, 600)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
  scale_linetype_manual(values = c("solid", "dotted", "dashed"), 
                        labels = c("Visit 1", "Visit 2", "6-Month")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_rt
```

## BPRS (Pt/Ctl)
### Total
```{r}
# run the LMER, report as an ANOVA
mod_bprs <- lmer(bprs_total ~ group * visit + (1|subj), data = dat_bprs)
anova_bprs <- anova(mod_bprs, type = 3)
anova_bprs

# run post-hoc tests on significant effects
print(emmeans(mod_bprs, list(pairwise ~ group), adjust = "mvt"))
```

### Positive
```{r}
# run the LMER, report as an ANOVA
mod_pos <- lmer(bprs_pos_ws ~ group * visit + (1|subj), data = dat_bprs)
anova_pos <- anova(mod_pos, type = 3)
anova_pos

# run post-hoc tests on significant effects
print(emmeans(mod_pos, list(pairwise ~ group), adjust = "mvt"))
```

### Negative
```{r}
# run the LMER, report as an ANOVA
mod_neg <- lmer(bprs_neg_ws ~ group * visit + (1|subj), data = dat_bprs)
anova_neg <- anova(mod_neg, type = 3)
anova_neg

# run post-hoc tests on significant effects
print(emmeans(mod_neg, list(pairwise ~ group), adjust = "mvt"))
print(emmeans(mod_neg, list(pairwise ~ visit), adjust = "mvt"))
print(emmeans(mod_neg, list(pairwise ~ group|visit), adjust = "mvt"))

### plot BPRS negative ----------
plot_neg <- ggplot(dat_bprs, aes(x= visit, y=bprs_neg_ws)) +
  stat_summary(fun="mean", geom="line", aes(color = group), linewidth = 1.5, position = position_dodge(0.2)) +
  stat_summary(fun="mean", geom="point", aes(color = group), size = 3, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BPRS Negative", 
       y = "BPRS Negative", 
       x = "Visit",
       color = "Group") +
  coord_cartesian(ylim = c(1, 2.5)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
plot_neg
```

### Disorg
```{r}
# run the LMER, report as an ANOVA
mod_disorg <- lmer(bprs_disorg_ws ~ group * visit + (1|subj), data = dat_bprs)
anova_disorg <- anova(mod_disorg, type = 3)
anova_disorg

# run post-hoc tests on significant effects
print(emmeans(mod_disorg, list(pairwise ~ group), adjust = "mvt"))
```

### Mania
```{r}
# run the LMER, report as an ANOVA
mod_mania <- lmer(bprs_mania_ws ~ group * visit + (1|subj), data = dat_bprs)
anova_mania <- anova(mod_mania, type = 3)
anova_mania

# run post-hoc tests on significant effects
print(emmeans(mod_mania, list(pairwise ~ group), adjust = "mvt"))
```

### Depr-Anx
```{r}
# run the LMER, report as an ANOVA
mod_dep <- lmer(bprs_dep_anx_ws ~ group * visit + (1|subj), data = dat_bprs)
anova_dep <- anova(mod_dep, type = 3)
anova_dep

# run post-hoc tests on significant effects
print(emmeans(mod_dep, list(pairwise ~ group), adjust = "mvt"))
```

### Plot All BPRS
#### Prep data for plot
```{r}
# add aff/non-aff dx variable to BPRS
dat_bprs <- dat_bprs %>%
  left_join(
    dat_demo %>% select(subj, dx_affect),
    by = "subj"
  )

dat_bprs$dx_affect <- factor(dat_bprs$dx_affect,
                             levels = c(0, 1, 2),
                             labels = c("Ctl", "Non-Aff", "Aff"))

# convert BPRS from wide to long format
dat_bprs_long <- dat_bprs %>% 
  # filter data to only keep variables of interest
  select(c(subj, visit, group, group_num, dx_affect, bprs_pos_ws, bprs_neg_ws,
           bprs_disorg_ws, bprs_mania_ws, bprs_dep_anx_ws)) %>% 
  # convert data to long format
  pivot_longer(cols = c(bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws, 
                        bprs_mania_ws, bprs_dep_anx_ws), 
               names_to = "bprs_symptom", 
               values_to = "bprs_score") %>% 
  # create labels
  mutate(bprs_symptom = dplyr::recode(bprs_symptom,
         "bprs_pos_ws" = "Pos",
         "bprs_neg_ws" = "Neg",
         "bprs_disorg_ws" = "Disorg",
         "bprs_mania_ws" = "Mania",
         "bprs_dep_anx_ws" = "Dep-Anx"),
          # set labels to factors, to appear in specified order
         bprs_symptom = factor(bprs_symptom,
                               levels = c("Pos", "Neg", "Disorg", 
                                          "Mania", "Dep-Anx"))) %>% 
  mutate(group_visit = factor(paste(group, visit, sep = "_"),
                              levels = c("Control_BL1", "Control_6M", 
                                         "Patient_BL1", "Patient_6M"))) %>% 
  mutate(visit = factor(visit,
                        levels = c("BL1", "6M")))
```

#### Line plot
```{r}
# Line plot of BPRS all symptoms
plot_bprs <- ggplot(dat_bprs_long, 
                    aes(x= bprs_symptom, y=bprs_score, 
                            group = interaction(group,visit))) +
  stat_summary(fun="mean", geom="line",
               aes(color = group, linetype = visit),
               linewidth = 1, position = position_dodge(0.2)) +
  stat_summary(fun="mean", geom="point", aes(color = group), 
               size = 2, position = position_dodge(0.2)) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = group), 
               width = 0.7, size = 1, position = position_dodge(0.2)) + 
  labs(title = "BPRS Symptoms", 
       y = "Average Item Score", 
       x = "Symptom Subfactors", 
       color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(1, NA)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Ctl", "EP")) + 
  scale_linetype_manual(values = c("solid", "dashed"), 
                        labels = c("Visit 1", "6-Month")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 12),
        axis.title = element_text(face="bold", size = 16),
        legend.text = element_text(face="bold", size = 12),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 20))
plot_bprs

# save plot to computer/local device
ggsave(paste0("graphs/", format(Sys.Date(), "%Y%m%d"), 
              "_BPRS_line_longitudinal.png"),
         plot_bprs, width = 6, height = 4)
```

#### Bar plot 1
```{r}
# Bar plot of BPRS all symptom
plot_bprs_bar <- ggplot(
  dat_bprs_long,
  aes(x = bprs_symptom, y = bprs_score, fill = group_visit)
) +
  stat_summary(
    fun = "mean",
    geom = "bar",
    position = position_dodge(width = 0.9),
    color = "black"
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    position = position_dodge(width = 0.9),
    width = 0.3
  ) +
  labs(
    title = "BPRS Symptoms",
    y = "Average Item Score", 
    x = "Symptom Subfactors", 
    fill = "Group x Visit"
  ) +
  scale_fill_manual(
    values = c(
      "Control_BL1" = "#2C737F",
      "Control_6M" = "#6BAEB7",
      "Patient_BL1"  = "#AA8829",
      "Patient_6M"  = "#D4B46A"),
    labels = c("Ctl Visit 1", "Ctl 6-Month", "EP Visit 1", "EP 6-Month")
  ) +
  coord_cartesian(ylim = c(1, NA)) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 16),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 16),
    plot.title = element_text(face = "bold", size = 22)
  )
plot_bprs_bar

# save plot to computer/local device
ggsave(paste0("graphs/", format(Sys.Date(), "%Y%m%d"), 
              "_BPRS_bar_longitudinal.png"),
         plot_bprs_bar, width = 6, height = 4)
```

#### Bar plot 2
```{r}
plot_bprs_bar2 <- ggplot(
  dat_bprs_long,
  aes(
    x = bprs_symptom,
    y = bprs_score,
    fill = group,
    pattern = visit
  )
) +
  stat_summary(
    fun = "mean",
    geom = "bar_pattern",
    position = position_dodge(width = 0.9),
    color = "black",
    pattern_fill = "black",
    pattern_color = "black",
    pattern_density = 0.1,
    pattern_spacing = 0.02
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    position = position_dodge(width = 0.9),
    width = 0.3
  ) +
  coord_cartesian(ylim = c(1, NA)) +
  labs(
    title = "BPRS Symptoms",
    y = "Average Item Score",
    x = "Symptom Subfactors", 
    fill = "Group",
    pattern = "Visit"
  ) +
  scale_fill_manual(
    values = c("Control" = "#2C737F", "Patient" = "#AA8829"),
    labels = c("Ctl", "EP")
  ) +
  scale_pattern_manual(
    values = c("BL" = "none", "6M" = "stripe"),
    labels = c("Visit 1", "6-Month")
  ) +
  guides(
  fill = guide_legend(override.aes = list(pattern = "none")),
  pattern = guide_legend()
) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 16),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 16),
    plot.title = element_text(face = "bold", size = 22)
  )
plot_bprs_bar2

# save plot to computer/local device
ggsave(paste0("graphs/", format(Sys.Date(), "%Y%m%d"), 
              "_BPRS_bar2_longitudinal.png"),
         plot_bprs_bar2, width = 6, height = 4)
```

## BPRS (Aff/NonAff)
### Total
```{r}
# run the LMER, report as an ANOVA
mod_bprs2 <- lmer(bprs_total ~ dx_affect * visit + (1|subj), data = dat_bprs)
anova_bprs2 <- anova(mod_bprs2, type = 3)
anova_bprs2

# run post-hoc tests on significant effects
print(emmeans(mod_bprs2, list(pairwise ~ dx_affect), adjust = "mvt"))
```

### Positive
```{r}
# run the LMER, report as an ANOVA
mod_pos2 <- lmer(bprs_pos_ws ~ dx_affect * visit + (1|subj), data = dat_bprs)
anova_pos2 <- anova(mod_pos2, type = 3)
anova_pos2

# run post-hoc tests on significant effects
print(emmeans(mod_pos2, list(pairwise ~ dx_affect), adjust = "mvt"))
```

### Negative
```{r}
# run the LMER, report as an ANOVA
mod_neg2 <- lmer(bprs_neg_ws ~ dx_affect * visit + (1|subj), data = dat_bprs)
anova_neg2 <- anova(mod_neg2, type = 3)
anova_neg2

# run post-hoc tests on significant effects
print(emmeans(mod_neg2, list(pairwise ~ dx_affect), adjust = "mvt"))
print(emmeans(mod_neg2, list(pairwise ~ visit), adjust = "mvt"))
print(emmeans(mod_neg2, list(pairwise ~ dx_affect|visit), adjust = "mvt"))

### plot BPRS negative ----------
# plot_neg2 <- ggplot(dat_bprs, aes(x= visit, y=bprs_neg_ws)) +
#   stat_summary(fun="mean", geom="line", aes(color = dx_affect, group = dx_affect), linewidth = 1.5, position = position_dodge(0.2)) +
#   stat_summary(fun="mean", geom="point", aes(color = dx_affect), size = 3, position = position_dodge(0.2)) + 
#   stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = dx_affect), 
#                width = 0.7, size = 1, position = position_dodge(0.2)) + 
#   labs(title = "BPRS Negative", 
#        y = "BPRS Negative", 
#        x = "Visit",
#        color = "Group") +
#   coord_cartesian(ylim = c(1, NA)) +
#   scale_color_manual(values=c("#2C737F", "#AA8829", "#C44E52"),
#                      labels=c("Control", "Non-Affective", "Affective")) + 
#   theme(panel.grid.major = element_blank(), 
#         panel.grid.minor = element_blank(),
#         panel.background = element_blank(), 
#         axis.line = element_line(colour = "black"), 
#         axis.text = element_text(face="bold", size = 16),
#         axis.title = element_text(face="bold", size = 20),
#         legend.text = element_text(face="bold", size = 14),
#         legend.title = element_text(face = "bold", size = 16),
#         plot.title = element_text(face = "bold", size = 22))
# plot_neg2
```

### Disorg
```{r}
# run the LMER, report as an ANOVA
mod_disorg2 <- lmer(bprs_disorg_ws ~ dx_affect * visit + (1|subj), data = dat_bprs)
anova_disorg2 <- anova(mod_disorg2, type = 3)
anova_disorg2

# run post-hoc tests on significant effects
print(emmeans(mod_disorg2, list(pairwise ~ dx_affect), adjust = "mvt"))
print(emmeans(mod_disorg2, list(pairwise ~ dx_affect|visit), adjust = "mvt"))

```

### Mania
```{r}
# run the LMER, report as an ANOVA
mod_mania <- lmer(bprs_mania_ws ~ dx_affect * visit + (1|subj), data = dat_bprs)
anova_mania <- anova(mod_mania, type = 3)
anova_mania

# run post-hoc tests on significant effects
print(emmeans(mod_mania, list(pairwise ~ dx_affect), adjust = "mvt"))
```

### Depr-Anx
```{r}
# run the LMER, report as an ANOVA
mod_dep <- lmer(bprs_dep_anx_ws ~ dx_affect * visit + (1|subj), data = dat_bprs)
anova_dep <- anova(mod_dep, type = 3)
anova_dep

# run post-hoc tests on significant effects
print(emmeans(mod_dep, list(pairwise ~ dx_affect), adjust = "mvt"))
```


### Bar plot
```{r}
# as needed, run previous section "prep data for plot" to create dat_bprs_long

# modify affect variable
dat_bprs_long <- dat_bprs_long %>% 
  mutate(affect = dx_affect) %>% 
  mutate(affect_num = dplyr::recode(affect,
         "Ctl" = 0,
         "Non-Aff" = 1,
         "Aff" = 2)) %>% 
  mutate(affect_visit = factor(paste(affect, visit, sep = "_"),
                              levels = c("Ctl_BL1", "Ctl_6M", 
                                         "Non-Aff_BL1", "Non-Aff_6M",
                                         "Aff_BL1", "Aff_6M")))


# Bar plot of BPRS all symptom
plot_bprs_bar3 <- ggplot(
  dat_bprs_long,
  aes(x = bprs_symptom, y = bprs_score, fill = affect_visit)
) +
  stat_summary(
    fun = "mean",
    geom = "bar",
    position = position_dodge(width = 0.9),
    color = "black"
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    position = position_dodge(width = 0.9),
    width = 0.3
  ) +
  labs(
    title = "BPRS Symptoms",
    y = "Average Item Score", 
    x = "Symptom Subfactors", 
    fill = "Group x Visit"
  ) +
  scale_fill_manual(
    values = c(
      "Ctl_BL1" = "#2C737F",
      "Ctl_6M" = "#6BAEB7",
      "Non-Aff_BL1"  = "#AA8829",
      "Non-Aff_6M"  = "#D4B46A",
      "Aff_BL1" = "#C44E52",
      "Aff_6M"  = "#d98d8f"),
    labels = c("Ctl Visit 1", "Ctl 6-Month", "Non-Aff Visit 1", "Non-Aff 6-Month",
               "Aff Visit 1", "Aff 6-Month")
  ) +
  coord_cartesian(ylim = c(1, NA)) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 16),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 16),
    plot.title = element_text(face = "bold", size = 22)
  )
plot_bprs_bar3

# save plot to computer/local device
ggsave(paste0("graphs/", format(Sys.Date(), "%Y%m%d"), 
              "_BPRS_affect_longitudinal.png"),
         plot_bprs_bar3, width = 6, height = 4)

# Ensure separate group and visit columns exist (split from affect_visit if needed)
dat_bprs_long <- dat_bprs_long %>%
  mutate(
    group = case_when(
      grepl("^Ctl", affect_visit) ~ "Ctl",
      grepl("^Non-Aff", affect_visit) ~ "Non-Aff",
      grepl("^Aff", affect_visit) ~ "Aff"
    ),
    visit = case_when(
      grepl("BL1$", affect_visit) ~ "BL1",
      grepl("6M$", affect_visit) ~ "6M"
    ),
    group = factor(group, levels = c("Ctl", "Non-Aff", "Aff")),
    visit = factor(visit, levels = c("BL1", "6M"))
  )


plot_bprs_bar4 <- ggplot(
  dat_bprs_long,
  aes(
    x = bprs_symptom,
    y = bprs_score,
    fill = group,
    pattern = visit
  )
) +
  stat_summary(
    fun = "mean",
    geom = "bar_pattern",
    position = position_dodge(width = 0.9),
    color = "black",
    pattern_fill = "black",
    pattern_color = "black",
    pattern_density = 0.1,
    pattern_spacing = 0.02
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    position = position_dodge(width = 0.9),
    width = 0.3
  ) +
  coord_cartesian(ylim = c(1, NA)) +
  labs(
    title = "BPRS Symptoms",
    y = "Average Item Score",
    x = "Symptom Subfactors",
    fill = "Group",
    pattern = "Visit"
  ) +
  scale_fill_manual(
    values = c("Ctl" = "#2C737F", "Non-Aff" = "#AA8829", "Aff" = "#C44E52"),
    labels = c("Ctl", "Non-Aff", "Aff")
  ) +
  scale_pattern_manual(
    values = c("BL1" = "none", "6M" = "stripe"),
    labels = c("Visit 1", "6-Month")
  ) +
  guides(
    fill = guide_legend(override.aes = list(pattern = "none")),
    pattern = guide_legend()
  ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 16),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 16),
    plot.title = element_text(face = "bold", size = 22)
  )
plot_bprs_bar4

# save plot to computer/local device
ggsave(paste0("graphs/", format(Sys.Date(), "%Y%m%d"), 
              "_BPRS_affect_v2_longitudinal.png"),
         plot_bprs_bar4, width = 6, height = 4)

```

# Global Function
## Social
```{r}
# run the LMER, report as an ANOVA
mod_gfs <- lmer(gfs ~ group * visit + (1|subj), data = dat_gf)
anova_gfs <- anova(mod_gfs, type = 3)
anova_gfs

# run post-hoc tests on significant effects
print(emmeans(mod_gfs, list(pairwise ~ group), adjust = "mvt"))
```

## Role
```{r}
# run the LMER, report as an ANOVA
mod_gfr <- lmer(gfr ~ group * visit + (1|subj), data = dat_gf)
anova_gfr <- anova(mod_gfr, type = 3)
anova_gfr

# run post-hoc tests on significant effects
print(emmeans(mod_gfr, list(pairwise ~ group), adjust = "mvt"))
```

## Bar plot
```{r}
# reshape data for plot
dat_gf_long <- dat_gf %>%
  select(subj, visit, group, gfs, gfr) %>%
  pivot_longer(
    cols = c(gfs, gfr),
    names_to = "gf_type",
    values_to = "gf_score"
  ) %>%
  mutate(
    gf_type = dplyr::recode(gf_type,
      "gfs" = "GFS",
      "gfr" = "GFR"
    ),
    gf_type = factor(gf_type, levels = c("GFS", "GFR")),
    visit = factor(visit, levels = c("BL1", "6M")),
    group_visit = factor(
      paste(group, visit, sep = "_"),
      levels = c("Control_BL1", "Control_6M", "Patient_BL1", "Patient_6M")
    ))

# plot it
plot_gf_bar <- ggplot(
  dat_gf_long,
  aes(x = gf_type, y = gf_score, fill = group_visit)
) +
  stat_summary(
    fun = "mean",
    geom = "bar",
    position = position_dodge(width = 0.9),
    color = "black"
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    position = position_dodge(width = 0.9),
    width = 0.3
  ) +
  labs(
    title = "Global Functioning",
    y = "Average Score",
    x = "Measure",
    fill = "Group x Visit"
  ) +
  scale_fill_manual(
    values = c(
      "Control_BL1" = "#2C737F",
      "Control_6M" = "#6BAEB7",
      "Patient_BL1" = "#AA8829",
      "Patient_6M" = "#D4B46A"
    ),
    labels = c("Ctl Visit 1", "Ctl 6-Month", "EP Visit 1", "EP 6-Month")
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    axis.line = element_line(colour = "black"),
    axis.text = element_text(face = "bold", size = 12),
    axis.title = element_text(face = "bold", size = 16),
    legend.text = element_text(face = "bold", size = 12),
    legend.title = element_text(face = "bold", size = 16),
    plot.title = element_text(face = "bold", size = 22)
  )

plot_gf_bar
```


# ICCs
## prep data for ICCs
```{r}
# remove any ppts who don't have both timepoints
dat_bprs_bl_6m <- dat_bprs %>%
  group_by(subj) %>%
  filter(all(c("BL1", "6M") %in% visit)) %>%
  ungroup()

# remove BL2
dat_topx_bl_6m <- dat_topx %>%
  filter(visit != "BL2")

# remove ppts who don't have both timepoints
dat_topx_bl_6m <- dat_topx_bl_6m %>% 
  group_by(subj) %>%
  filter(all(c("BL1", "6M") %in% visit)) %>%
  ungroup()

# remove duplicate 6M visit
dat_topx_bl_6m <- dat_topx_bl_6m %>% 
  filter(!(subj == "SP2150" & visit == "6M" & date == "20231130"))
```

## function for TOPX ICC
```{r}
dat_icc_topx <- dat_topx_bl_6m %>% 
  select(c(subj, visit, trial_type, error))

# version 1 - done manually
# dat_icc_AX <- dat_icc_topx %>% 
#   filter(trial_type == "AX") %>%
#   pivot_wider(names_from = visit, 
#               values_from = error) %>%
#   select(-c(subj, trial_type))
# ICC(dat_icc_AX)
# 
# dat_icc_AY <- dat_icc_topx %>% 
#   filter(trial_type == "AY") %>%
#   pivot_wider(names_from = visit, 
#               values_from = error) %>%
#   select(-c(subj, trial_type))
# ICC(dat_icc_AY)
# 
# dat_icc_BX <- dat_icc_topx %>% 
#   filter(trial_type == "BX") %>%
#   pivot_wider(names_from = visit, 
#               values_from = error) %>%
#   select(-c(subj, trial_type))
# ICC(dat_icc_BX)
# 
# dat_icc_BY <- dat_icc_topx %>% 
#   filter(trial_type == "BY") %>%
#   pivot_wider(names_from = visit, 
#               values_from = error) %>%
#   select(-c(subj, trial_type))
# ICC(dat_icc_BY)

## version 2 - returns entire ICC table
# library(purrr)
# icc_results <- c("AX", "AY", "BX", "BY") %>%
#   set_names() %>%
#   map(~ dat_icc_topx %>%
#         filter(trial_type == .x) %>%
#         pivot_wider(names_from = visit, values_from = error) %>%
#         select(-subj, -trial_type) %>%
#         ICC())
# icc_results

## version 3 - provides the whole table, doesn't sort by group
calc_ICC_topx <- function(icc_trialtype, var = "error") {
  library(dplyr)
  library(tidyr)
  library(psych)
  dat_icc <- dat_topx_bl_6m %>%
    filter(trial_type == icc_trialtype) %>%
    select(subj, visit, !!sym(var)) %>%
    pivot_wider(names_from = visit, values_from = !!sym(var)) %>%
    select(-subj)
  icc_out <- ICC(dat_icc, lmer = TRUE)
  return(icc_out)
}
calc_ICC_topx("AX", "error")
calc_ICC_topx("AY", "error")
calc_ICC_topx("BX", "error")
calc_ICC_topx("BY", "error")
```

## function for BPRS ICC
```{r}
dat_icc_bprs <- dat_bprs_bl_6m %>% 
  select(c(subj, visit, bprs_pos_ws, 
           bprs_neg_ws, bprs_disorg_ws, bprs_mania_ws,
           bprs_dep_anx_ws)) %>% 
  pivot_longer(cols = c(bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws,
                        bprs_mania_ws, bprs_dep_anx_ws),
             names_to = "bprs", values_to = "value") %>% 
  mutate(bprs = dplyr::recode(bprs,
                              bprs_pos_ws = "pos",
                              bprs_neg_ws = "neg",
                              bprs_disorg_ws = "disorg",
                              bprs_mania_ws = "mania",
                              bprs_dep_anx_ws = "dep_anx"))

## version 3 - provides the whole table, doesn't sort by group
calc_ICC_bprs <- function(icc_bprsfactor, var = "value") {
  library(dplyr)
  library(tidyr)
  library(psych)
  dat_icc <- dat_icc_bprs %>%
    filter(bprs == icc_bprsfactor) %>%
    select(subj, visit, !!sym(var)) %>%
    pivot_wider(names_from = visit, values_from = !!sym(var)) %>%
    select(-subj)
  icc_out <- ICC(dat_icc, lmer = TRUE)
  return(icc_out)
}
calc_ICC_bprs("pos", "value")
calc_ICC_bprs("neg", "value")
calc_ICC_bprs("disorg", "value")
calc_ICC_bprs("mania", "value")
calc_ICC_bprs("dep_anx", "value")

```


# Spearman CORRELATIONS
## TOPX + BPRS
### Create Dataframe
```{r}
# create new bprs dataframe with only necessary variables
dat_bprs_abbrev <- dat_bprs %>% 
  select(subj, visit, bprs_total, bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws, bprs_mania_ws, bprs_dep_anx_ws)

# create new topx dataframe, eliminating BL2
dat_topx_abbrev <- dat_topx %>% 
  filter(visit == "BL1" | visit == "6M") 
dat_topx_abbrev <- dat_topx_abbrev %>% 
  select(subj, visit, group, d_context, d_expect, trial_type, rt_mdn, error)
#dat_topx_abbrev$visit <- as.character(dplyr::recode(dat_topx_abbrev$visit, "BL1" = "BL"))
dat_topx_abbrev <- pivot_wider(dat_topx_abbrev, 
                               id_cols = c(subj, visit, group, 
                                           d_context, d_expect), 
                               values_from = c(rt_mdn, error), 
                               names_from = trial_type)

# combine topx & bprs data
dat_topx_bprs <- dat_topx_abbrev %>%
  left_join(dat_bprs_abbrev %>% 
              select(subj, visit, bprs_total, bprs_pos_ws, 
                     bprs_neg_ws, bprs_disorg_ws, bprs_mania_ws, 
                     bprs_dep_anx_ws),
    by = c("subj", "visit"))

# create dataframes for each visit
dat_topx_bprs_bl <- dat_topx_bprs %>% 
  filter(visit == "BL1")
dat_topx_bprs_6m <- dat_topx_bprs %>% 
  filter(visit == "6M")

# filter out ctls
dat_topx_bprs_pt <- dat_topx_bprs_bl %>% 
  filter(group == "Patient")
```

### Test Normality
```{r}
# Test normal distribution of variables
# if p < .05, variable is non-normal/skewed 
shapiro.test(dat_topx_bprs$error_AX)
shapiro.test(dat_topx_bprs$error_AY)
shapiro.test(dat_topx_bprs$error_BX)
shapiro.test(dat_topx_bprs$error_BY)
shapiro.test(dat_topx_bprs$bprs_pos_ws)
shapiro.test(dat_topx_bprs$bprs_neg_ws)
shapiro.test(dat_topx_bprs$bprs_disorg_ws)
```

### Run correlation
```{r}
# create dataframe for a correlation matrix of TOPX error rates and BPRS symptoms
dat_cor1 <- dat_topx_bprs_pt %>% 
  select(error_AX, error_AY, error_BX, bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws)

# compute and print correlations + p-values
cor1_results <- rcorr(as.matrix(dat_cor1), type = "spearman")
(cor1_matrix <- round(cor1_results$r, 4))
(p1_matrix <- round(cor1_results$P, 4))
```

### Plot corr matrix
```{r}
# Plot Spearman correlation matrix with significance masking
filename <- paste0("graphs/", format(Sys.Date(), "%Y%m%d"), "_TOPX_BPRS.png")
col_pal <- colorRampPalette(c("#2C737F", "white", "#A30021"))(100)
colnames(cor1_matrix) <- c("AX Error", "AY Error", "BX Error",
                           "BPRS Positive", "BPRS Negative", "BPRS Disorganized")
rownames(cor1_matrix) <- c("AX Error", "AY Error", "BX Error",
                           "BPRS Positive", "BPRS Negative", "BPRS Disorganized")

# create plot as a function
plot_topx_bprs <- function() {
  corrplot(
    cor1_matrix,
    method = "color",
    col = col_pal,
    is.corr = FALSE,
    col.lim = range(cor1_matrix, na.rm = TRUE),
    type = "lower",
    diag = TRUE,
    
    # create asterisks for significance - not using because overlaps with coeff
    # p.mat = p1_matrix,
    # sig.level = 0.05,
    # insig = "label_sig",
    # pch.cex = 0.8,
    insig = "blank",
    
    addCoef.col = "black",
    number.cex = 0.7,
    
    tl.col = "black",
    tl.pos = "t",
    tl.cex = 0.7,
    tl.srt = 45,
    
    cl.pos = "r",
    cl.cex = 0.7,
    cl.ratio = 0.2,
    cl.align.text = "l",
    
    title = "Correlations of TOPX and BPRS at Visit 1",
    cex.main = .9,
    mar = c(1, 1, 2, 1),
    frame.plot = TRUE
  )
}

# save plot
filename <- paste0("graphs/", format(Sys.Date(), "%Y%m%d"), "_TOPX_BPRS.png")
png(filename, width = 6, height = 4, units = "in", res = 300)
plot_topx_bprs()
dev.off()

# print plot
plot_topx_bprs()
```

## BPRS + GF
### Create Dataframe
```{r}
# create new bprs dataframe with only necessary variables - already done above
# dat_bprs_abbrev <- dat_bprs %>% 
#   select(subj, visit, bprs_total, bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws, bprs_mania_ws, bprs_dep_anx_ws)

# create new topx dataframe, eliminating BL2
dat_gf_abbrev <- dat_gf %>% 
  select(subj, visit, group, gfs, gfr)

# combine topx & bprs data
dat_bprs_gf <- dat_bprs_abbrev %>%
  left_join(dat_gf_abbrev %>% 
              select(subj, visit, group, gfs, gfr),
    by = c("subj", "visit"))

# create dataframes for each visit
dat_bprs_gf_bl <- dat_bprs_gf %>% 
  filter(visit == "BL1")
dat_bprs_gf_6m <- dat_bprs_gf %>% 
  filter(visit == "6M")
dat_bprs_gf_pt <- dat_bprs_gf_bl %>% 
  filter(group == "Patient")
```

### Run correlation
```{r}
# create dataframe for a correlation matrix of TOPX error rates and BPRS symptoms
dat_cor2 <- dat_bprs_gf_pt %>% 
  select(bprs_pos_ws, bprs_neg_ws, bprs_disorg_ws, bprs_mania_ws, bprs_dep_anx_ws, gfs, gfr)

# compute and print correlations + p-values
cor2_results <- rcorr(as.matrix(dat_cor2), type = "spearman")
(cor2_matrix <- round(cor2_results$r,4))
(p2_matrix   <- round(cor2_results$P,4))
```

### Plot corr matrix
```{r}
# Plot Spearman correlation matrix with significance masking
filename <- paste0("graphs/", format(Sys.Date(), "%Y%m%d"), "_BPRS_GF.png")
col_pal <- colorRampPalette(c("#2C737F", "white", "#A30021"))(100)
colnames(cor2_matrix) <- c("BPRS Positive", "BPRS Negative", "BPRS Disorganized",
                           "BPRS Mania", "BPRS Depression-Anxiety", "Global Function: Social",
                           "Global Function: Role")
rownames(cor2_matrix) <- c("BPRS Positive", "BPRS Negative", "BPRS Disorganized",
                           "BPRS Mania", "BPRS Depression-Anxiety", "Global Function: Social",
                           "Global Function: Role")

# create plot as a function
plot_bprs_gf <- function() {
  corrplot(
    cor2_matrix,
    method = "color",
    col = col_pal,
    is.corr = FALSE,
    col.lim = range(cor2_matrix, na.rm = TRUE),
    type = "lower",
    diag = TRUE,
    
    # create asterisks for significance - not using because overlaps with coeff
    # p.mat = p1_matrix,
    # sig.level = 0.05,
    # insig = "label_sig",
    # pch.cex = 0.8,
    insig = "blank",
    
    addCoef.col = "black",
    number.cex = 0.7,
    
    tl.col = "black",
    tl.pos = "t",
    tl.cex = 0.7,
    tl.srt = 45,
    
    cl.pos = "r",
    cl.cex = 0.7,
    cl.ratio = 0.2,
    cl.align.text = "l",
    
    title = "Correlations of BPRS and GF at Visit 1",
    cex.main = .9,
    mar = c(1, 1, 2, 1),
    frame.plot = TRUE
  )
}

# save plot
filename <- paste0("graphs/", format(Sys.Date(), "%Y%m%d"), "_TOPX_BPRS.png")
png(filename, width = 6, height = 4, units = "in", res = 300)
plot_topx_bprs()
dev.off()

# print plot
plot_bprs_gf()
```
