---
title: "step_combine_behav_dat"
author: "Jessica Arend"
date: "2025-03-17"
output: html_document
---
# Load, clean, combine data [hidden in output]
```{r, echo = FALSE, include = FALSE}
# clear workspace
rm(list=ls())

# read in packages
pacman::p_load(stats, stringr, dplyr, tidyr, ggplot2, lme4, emmeans, lmerTest, Matrix, pbkrtest, psych, car, performance, parameters, data.table, tibble)

# set working directory and read in data
setwd("~/Documents/umn_work/topx_analyses/")
```

## Functions - cleaning
```{r, echo = FALSE, include = FALSE}
## function to remove test cases, fix name formatting -------
clean_data <- function(data) {
  # rename subject id as "subj"
  data <- data %>% rename(subj = record_id)
  # identify test rows and filter them out
  ex_strings <- c('TEST', "Test", "test", "NDAR", "archive")
  dat_probs <- data[grep(paste(ex_strings, collapse='|'), 
                         data$subj, ignore.case = TRUE), ]
  data <- data[!data$subj %in% dat_probs$subj, ]
  # remove the "proofpoint" string from subj ids
  data$subj <- gsub('PROOFPOINT|-|[*]','', data$subj)
  # trim white spaces
  data$subj <- trimws(data$subj)
  return(as.data.frame(data))
  }

# function to clean up blankrows and repeat instances
clean_rep_inst <- function(data, date) {
  # remove rows with all NAs
  data <- data[!is.na(data$redcap_repeat_instance), ]
  # if a given subj has multiple instances, keep 2nd instance
  data <- as.data.table(data)
  tmp <- data[order(data[[date]]), 
                      if (any(redcap_repeat_instance == "2")) 
                        first(.I[redcap_repeat_instance == "2"]) 
                      else last(.I), keyby = "subj"]
  data <- data[tmp$V1]
  return(as.data.frame(data))
}
```

## Demo
```{r, echo = FALSE, include = FALSE}
### demographics -------
# pull data from REDCap > "Demographics + Dx" report 
# load in and clean data
dat_demo <- read.csv("../data/nontask_data/demographics/step_demo_dx_20250916.csv")
dat_demo <- clean_data(dat_demo)

# remove unnecessary rows
dat_demo <- subset(dat_demo, select = -c(redcap_event_name, redcap_repeat_instrument, redcap_repeat_instance, clinical_status))

# recode gender into one variable, rather than two
dat_demo <- dat_demo %>% 
  mutate(gender = if_else(is.na(dat_demo$gender_cis), dat_demo$gender_trans, dat_demo$gender_cis),
                  .after = "gender_other")

# fix missing gender variables for nonbinary ppts
# manually code some nonbinary ppts
tnb_ppts <- c("SP1030", "SP1149", "SP2014", "SP2084", "SP2112", "SP2116", "SP2120", 
              "SP2128", "SP2144", "SP2149","SP2160", "SP2195", "SP2197", "SP2207")
dat_demo <- dat_demo %>% 
  mutate(gender = ifelse(subj %in% tnb_ppts, "3", gender))

# recode to 5-way split: 1=cis man, 2=cis woman, 3=trans man, 4=trans woman, 5 = enby
dat_demo <- dat_demo %>%
  mutate(
    gender_5split = case_when(
      cis_or_trans==1 & gender_cis==1 ~ 1, # cis man
      cis_or_trans==1 & gender_cis==2 ~ 2, # cis woman
      cis_or_trans==2 & gender_trans==1 ~ 3, #trans man
      cis_or_trans==2 & gender_trans==2 ~ 4, # trans woman
      cis_or_trans==2 & gender_trans==3 ~ 5), #nonbinary
    .after = "gender")
dat_demo <- dat_demo %>% 
  mutate(gender_5split = ifelse(subj %in% tnb_ppts, "5", gender_5split))

# recode cis or trans: 1 = cis man, 2 = cis woman, 3 = trans/enby
dat_demo <- dat_demo %>% 
  mutate(
    gender_3split = case_when(
      gender_5split == 1 ~ 1,
      gender_5split == 2 ~ 2, 
      gender_5split == 3 ~ 3,
      gender_5split == 4 ~ 3,
      gender_5split == 5 ~ 3),
    .after = "gender_5split")

# recode cis or trans: 0 = cis, 1 = trans/enby
dat_demo <- dat_demo %>% 
  mutate(
    gender_cis_trans = case_when(
      gender_5split == 1 ~ 0,
      gender_5split == 2 ~ 0, 
      gender_5split == 3 ~ 1,
      gender_5split == 4 ~ 1,
      gender_5split == 5 ~ 1),
    .after = "gender_5split")
dat_demo <- dat_demo %>% 
  mutate(gender_5split = ifelse(subj %in% tnb_ppts, "5", gender_5split))

# recode sexual orientation 1: heterosexual, 2 = non-het, 3 = UNK
dat_demo <- dat_demo %>%
  mutate(
    sexuality_3split = case_when(
      sexual_orientation == 1 ~ 1, # heterosexual
      sexual_orientation == 2 ~ 2, # gay or lesbian
      sexual_orientation == 3 ~ 2, # bisexual
      sexual_orientation == 4 ~ 3, # prefer not to say
      sexual_orientation == 5 ~ 2), # other
    .after = "sexual_orientation")

# recode lgbtq+
# gender_3split: 1 = cis man, 2 = cis woman, 3 = trans/enby
# recode sexual orientation 1: heterosexual, 2 = non-het, 3 = UNK
# lgbtq: 1 = cis-straight, 2 = queer/trans, 3 = UNK
dat_demo <- dat_demo %>%
  mutate(
    lgbtq = case_when(
      gender_3split==1 & sexuality_3split==1 ~ 1, # cis man-straight
      gender_3split==1 & sexuality_3split==2 ~ 2, # cis man-queer
      gender_3split==1 & sexuality_3split==3 ~ 3, # cis man-UNK
      gender_3split==2 & sexuality_3split==1 ~ 1, # cis woman-straight
      gender_3split==2 & sexuality_3split==2 ~ 2, # cis woman-queer
      gender_3split==2 & sexuality_3split==3 ~ 3, # cis woman-UNK
      gender_3split==3 & sexuality_3split==1 ~ 2, # trans-straight
      gender_3split==3 & sexuality_3split==2 ~ 2, # trans-queer
      gender_3split==3 & sexuality_3split==3 ~ 2), # trans-UNK
    .after = "sexuality_3split")

# recode race NAs/other
dat_demo <- dat_demo %>%
  mutate(race = ifelse(is.na(race), 6, race))
dat_demo <- dat_demo %>%
  mutate(race_5split = ifelse(race == 6, 5, race))

# fix missing hispanicity variables for nonbinary ppts
dat_demo <- dat_demo %>%
  mutate(hispanic = case_when(
    subj %in% c("SP1030") ~ 1,  # Recode to 1=yes
    TRUE ~ hispanic)) %>% 
  mutate(hispanic = case_when(
    subj %in% c("SP1097", "SP1099") ~ 0,
    TRUE ~ hispanic))


# recode "highest caregiver education" variable, taking highest of mother/father
dat_demo$hi_caregiver_education <- pmax(dat_demo$father_education, dat_demo$mother_education, na.rm = TRUE)

# recode status_ppt_type (numeric) to group (labelled category)
dat_demo$group <- ifelse(dat_demo$status_ppt_type == 1, "Control", 
                     ifelse(dat_demo$status_ppt_type == 2, 
                            "Patient", NA))

# if controls' diagnosis is NA, recode dx=0
dat_demo$primary_dx[dat_demo$status_ppt_type == 1 &
                  is.na(dat_demo$primary_dx)] <- 0

# set control duration of psychosis to NA
dat_demo$duration_of_psychosis[dat_demo$status_ppt_type == 1] <- NA

# recode affective/nonaffective psychosis
dat_demo$dx_affect <- as.character(dat_demo$primary_dx)
dat_demo$dx_affect <- as.numeric(
  dplyr::recode(dat_demo$dx_affect,
                "0" = "0", # control
                "1" = "1", # schizophrenia -> non-affective (NA)
                "2" = "1", # schizoaffective -> NA
                "3" = "1", # schizophreniform -> NA
                "4" = "1", # psychosis NOS -> NA
                "5" = "2", # bipolar with psychosis -> Aff
                "6" = "2", # major depression with psychosis - Aff
                "7" = "1")) # other unspecified psychosis -> NA

# recode diagnosis - 5 split
# before: 0=control, 1=Sz, 2=SzAff, 3=Szform, 4=NOS, 5=BP, 6=MDD, 7=other
# now: 0 = control, 1 = Sz/SzAff/Szform, 2 = BP, 3 = MDD, 4 = NOS/other
dat_demo$dx_5split <- as.character(dat_demo$primary_dx)
dat_demo$dx_5split <- as.numeric(
  dplyr::recode(dat_demo$dx_5split,
                "0" = "0",
                "1" = "1",
                "2" = "1",
                "3" = "1",
                "4" = "4",
                "5" = "2",
                "6" = "3",
                "7" = "4"))

# filter out if BL2 not complete
dat_demo <- subset(dat_demo, b2_status == "4"| b2_status == "6") # completed/withdrew at baseline 2
# dat_demo <- subset(dat_demo, fu6_status == "4") # completed 6mo

# start the revised combined data frame
dat_behav <- dat_demo
```

## optional: filter dat behav by another dataset
```{r}
ddm_sample_subjx <- read.csv("~/Documents/umn_work/topx_analyses/topx_ddm/data/topx_ddm_sample_subjx_20250316.csv")
dat_behav <- subset(dat_behav, subj %in% ddm_sample_subjx$subj)
```

## BPRS
```{r, echo = FALSE, include = FALSE}
### BPRS  -------------------
# download BPRS data from REDCap using "Data Exports" > "Selected Instruments or Events" >
    #  select "BPRS" & "Baseline" > export raw data as csv
# load in and clean data
dat_bprs <- read.csv("../data/nontask_data/clinical/bprs_data_20250317.csv")
dat_bprs <- clean_data(dat_bprs)
dat_bprs <- clean_rep_inst(dat_bprs, "bprs_date")
dat_bprs <- dat_bprs %>%
  mutate(across(c(bprs02_anxi, bprs03_depr, bprs14_diso, bprs18_motr, 
                  bprs19_tens, bprs22_distr, bprs23_mohy, bprs24_mann), 
                as.integer))

dat_bprs <- dat_bprs %>% 
  mutate(bprs_pos_ws = bprs08_gran + bprs09_susp + bprs10_hall + bprs11_unus + bprs12_bizb) %>%
  mutate(bprs_neg_ws = bprs16_blun + bprs17_emot + bprs18_motr) %>% 
  mutate(bprs_disorg_ws = bprs01_somc + bprs06_host + bprs13_self + 
           bprs14_diso + bprs15_conc + bprs20_unco) %>% 
  mutate(bprs_mania_ws = bprs07_elat + bprs08_gran + bprs21_exci + bprs23_mohy + bprs24_mann) %>% 
  mutate(bprs_dep_anx_ws = bprs02_anxi + bprs03_depr + bprs04_suic + bprs05_guil)

dat_bprs <- dat_bprs %>%
  select(
    subj,
    bprs_repeat_instance = redcap_repeat_instance,
    bprs_date,
    bprs_total,
    bprs_positive,
    bprs_negative,
    bprs_disorganized,
    bprs_depression,
    bprs_mania,
    bprs_pos_ws,
    bprs_neg_ws,
    bprs_disorg_ws,
    bprs_mania_ws,
    bprs_dep_anx_ws
  )

# match BPRS data to all behav data
dat_behav <- merge(dat_behav, dat_bprs, by = "subj", all.x = TRUE)
```
 
## SANS-SAPS
```{r, echo = FALSE, include = FALSE}
### SANS-SAPS  ---------------------
# download SANS-SAPS data from REDCap using "Data Exports" > "Selected Instruments or Events" >
    #  select "SANS-SAPS" & "Baseline" > export raw data as csv
# load in and clean data
dat_sanssaps <- read.csv("../data/nontask_data/clinical/step_sanssaps_20250125.csv")
dat_sanssaps <- clean_data(dat_sanssaps)
dat_sanssaps <- clean_rep_inst(dat_sanssaps, "sanssaps_date")
dat_sanssaps <- dat_sanssaps %>% filter(!grepl("^SP1", subj))

# select only columns with rated (numeric) values
dat_sanssaps <- dat_sanssaps %>% 
  select(subj, sanssaps_date, redcap_repeat_instance, sans01_express, sans02_movement, sans03_gestures, sans04_eye, sans05_respons, sans06_vocal, sans07_global_flattening, sans08_inapprop, sans09_povspeech, sans10_povcontent, sans11_blocking, sans12_latency, sans13_global_alogia, sans14_hygiene, sans15_impersist, sans16_anergia, sans17_global_avolition, sans18_recreate, sans19_sexual, sans20_intimacy, sans21_friends, sans22_global_asociality, sans23_socinattention, sans24_statinattention, sans25_global_attention, saps01_auditory, saps02_comment, saps03_converse, saps04_somatic, saps05_olfactory, saps06_visual, saps07_global_hall, saps08_persec, saps09_jealous, saps10_guilt, saps11_grand, saps12_religious, saps13_somaticdel, saps14_reference, saps15_control, saps16_mind, saps17_broadcast, saps18_insertion, saps19_withdrawal, saps20_global_del, saps21_appearance, saps22_social, saps23_aggress, saps24_repetitive, saps25_global_biz, saps26_derailment, saps27_tangent, saps28_incoherence, saps29_illogicality, saps30_circumstan, saps31_pressured, saps32_distract, saps33_clanging, saps34_global_thought)

# make all sans/saps items numeric
columns_to_convert <- c("sans01_express", "sans02_movement", "sans03_gestures", "sans04_eye", "sans05_respons", "sans06_vocal", "sans08_inapprop", "sans09_povspeech", "sans10_povcontent", "sans11_blocking", "sans12_latency", "sans14_hygiene", "sans15_impersist", "sans16_anergia", "sans18_recreate", "sans19_sexual", "sans24_statinattention", "saps21_appearance", "saps24_repetitive")
dat_sanssaps <- dat_sanssaps %>%
  mutate_at(vars(columns_to_convert), as.integer)


# note: sans08 is typically not included in sans item total; saps08/saps10/saps13 not typically included in saps item total
dat_sanssaps <- dat_sanssaps %>% 
  mutate(sans_item_total = sans01_express + sans02_movement +
           sans03_gestures + sans04_eye + sans05_respons +
           sans06_vocal + sans08_inapprop + sans09_povspeech +
           sans10_povcontent + sans11_blocking + sans12_latency +
           sans14_hygiene + sans15_impersist + sans16_anergia +
           sans18_recreate + sans19_sexual + sans20_intimacy +
           sans21_friends + sans23_socinattention +
           sans24_statinattention) %>%
  mutate(sans_global_total = sans07_global_flattening +
           sans13_global_alogia + sans17_global_avolition +
           sans22_global_asociality + sans25_global_attention) %>% 
  mutate(saps_item_total = saps01_auditory + saps02_comment + 
           saps03_converse + saps04_somatic +
           saps05_olfactory + saps06_visual + saps08_persec +
           saps09_jealous + saps10_guilt + saps11_grand +
           saps12_religious + saps13_somaticdel + saps14_reference +
           saps15_control + saps16_mind + saps17_broadcast +
           saps18_insertion + saps19_withdrawal + saps21_appearance +
           saps22_social + saps23_aggress + saps24_repetitive +
           saps26_derailment + saps27_tangent + saps28_incoherence +
           saps29_illogicality + saps30_circumstan +
           saps31_pressured + saps32_distract + saps33_clanging) %>%
  mutate(saps_global_total = saps07_global_hall + saps20_global_del +
           saps25_global_biz + saps34_global_thought) %>% 
  mutate(sanssaps_total = sans_item_total + saps_item_total)

# Tibber 1st order factors - positive (tht disorder, biz beh, grand/rel, oth biz)
dat_sanssaps <- dat_sanssaps %>% 
  mutate(sanssaps_tht_dis_tib1a = saps26_derailment +
           saps27_tangent + saps28_incoherence + saps29_illogicality +
           saps30_circumstan + saps31_pressured + saps32_distract +
           saps33_clanging + sans10_povcontent + sans24_statinattention) %>% 
  mutate(sanssaps_biz_beh_tib1a = saps22_social + saps23_aggress) %>% 
  mutate(sanssaps_grand_rel_tib1a = saps11_grand + saps12_religious +
           saps22_social) %>%
  mutate(sanssaps_oth_biz_tib1a = saps21_appearance + saps24_repetitive +
           saps28_incoherence + sans14_hygiene) %>% 
# Tibber 1st order factors - negative (neg, soc dys, alog/inattn)
# (1a includes all, 1b excludes overlapping)
  mutate(sanssaps_neg_tib1a = sans01_express + sans02_movement +
           sans03_gestures + sans04_eye + sans05_respons +
           sans06_vocal + sans09_povspeech + sans11_blocking +
           sans12_latency + sans16_anergia + sans21_friends) %>% 
  mutate(sanssaps_soc_dys_tib1a = sans14_hygiene +
           sans15_impersist + sans16_anergia + sans18_recreate +
           sans19_sexual + sans20_intimacy + sans21_friends 
         + sans23_socinattention) %>% 
  mutate(sanssaps_alog_inat_tib1a = sans09_povspeech +
           sans10_povcontent + sans11_blocking + sans12_latency +
           sans23_socinattention + sans24_statinattention) %>%
  mutate(sanssaps_neg_tib1b = sans01_express + sans02_movement +
           sans03_gestures + sans04_eye + sans05_respons +
           sans06_vocal + sans09_povspeech + sans12_latency) %>%
  mutate(sanssaps_soc_dys_tib1b = sans14_hygiene + sans15_impersist +
           sans16_anergia + sans18_recreate + sans19_sexual +
           sans20_intimacy + sans21_friends) %>% 
  mutate(sanssaps_alog_inat_tib1b = sans10_povcontent +
           sans11_blocking + sans23_socinattention +
           sans24_statinattention) %>% 
# Tibber 1st order factors - disorganization (delu, AH, other halluc)
  mutate(sanssaps_delu_tib1a = saps14_reference +
           saps15_control + saps16_mind + saps17_broadcast +
           saps18_insertion + saps19_withdrawal) %>% 
  mutate(sanssaps_aud_hall_tib1a = saps01_auditory +
           saps02_comment + saps03_converse + saps17_broadcast) %>% 
  mutate(sanssaps_oth_hall_tib1a = saps04_somatic +
           saps05_olfactory + saps06_visual) %>% 
# Tibber 2nd order factors - positive
  mutate(sanssaps_pos_tib2 = (sanssaps_tht_dis_tib1a +
                                sanssaps_biz_beh_tib1a +
                                sanssaps_grand_rel_tib1a +
                                sanssaps_oth_biz_tib1a)/19) %>% 
#Tibber 2nd order factors - negative
  mutate(sanssaps_neg_tib2 = (sanssaps_neg_tib1a +
                                sanssaps_soc_dys_tib1a +
                                sanssaps_alog_inat_tib1a)/25) %>% 
  mutate(sanssaps_neg_tib2b = round(((sanssaps_neg_tib1b +
                                        sanssaps_soc_dys_tib1b +
                                        sanssaps_alog_inat_tib1b)/19),2)) %>% 
# Tibber 2nd order factors - disorganized
  mutate(sanssaps_disorg_tib2 = (sanssaps_delu_tib1a +
                                   sanssaps_aud_hall_tib1a +
                                   sanssaps_oth_hall_tib1a)/13)
  
dat_sanssaps <- dat_sanssaps %>%
  select(
    subj,
    sanssaps_repeat_instance = redcap_repeat_instance,
    sanssaps_date,
    sans_global_total,
    sans_item_total,
    saps_global_total,
    saps_item_total,
    sanssaps_total,
    sanssaps_pos_tib2, 
    sanssaps_neg_tib2, 
    sanssaps_neg_tib2b, 
    sanssaps_disorg_tib2
  )

# match SANS-SAPS data to all behav data
dat_behav <- merge(dat_behav, dat_sanssaps, by = "subj", all.x = TRUE)
```

## SPQ
```{r, echo = FALSE, include = FALSE}
### SPQ  -------------------
# load in and clean data
dat_spq <- read.csv("../data/nontask_data/clinical/step_spqbr_20250125.csv")
dat_spq <- clean_data(dat_spq)
dat_spq <- clean_rep_inst(dat_spq, "spqbr_date")

# create dataframe with only target variables
dat_spq <- dat_spq %>%
  select(subj, spqbr_date, redcap_repeat_instance,
         spqbr01_talking:spqbr32_unus_large) 
colnames(dat_spq) <- c("subj", "spq_date", "spq_repeat_instance", paste0("spq_", 1:32))

# calculate subscales
dat_spq <- dat_spq %>% 
  mutate(
    spq_ir = spq_1 + spq_2 + spq_3, # ideas of reference
    spq_s = spq_4 + spq_5 + spq_6, # suspiciousness
    spq_cf = spq_7 + spq_8 + spq_9, # no close friends
    spq_ca = spq_10 + spq_11 + spq_12,# constricted affect
    spq_eb = spq_13 + spq_14 + spq_15 + spq_16, # eccentric behavior
    spq_sa = spq_17 + spq_18 + spq_19 + spq_20, # social anxiety
    spq_mt = spq_21 + spq_22 + spq_23 + spq_24, # magical thinking
    spq_os = spq_25 + spq_26 + spq_27 + spq_28, # odd speech
    spq_up = spq_29 + spq_30 + spq_31 + spq_32, # unusual perception
    # subscales:
    spq_cp = spq_ir + spq_s + spq_mt + spq_up, # cognitive perceptual
    spq_do = spq_eb + spq_os, # disorganized
    spq_ip = spq_cf + spq_ca + spq_sa, # interpersonal
    spq_total = spq_cp + spq_do + spq_ip) # total

# filter dataframe to only include total and subscales
dat_spq <- dat_spq %>%
  select(subj, spq_repeat_instance, spq_date, spq_total, spq_cp, spq_do, spq_ip)

# match SPQ data to all behav data
dat_behav <- merge(dat_behav, dat_spq, by = "subj", all.x = TRUE)
```

## GFS/GFR
```{r, echo = FALSE, include = FALSE}
### GFS/GFR  ------------------
# load in and clean data
dat_gf <- read.csv("../data/nontask_data/functioning/step_gfs-gfr_20250125.csv")
dat_gf <- clean_data(dat_gf)
dat_gf <- clean_rep_inst(dat_gf, "gbl_func_date")

# create dataframe with only target variables
dat_gf <- dat_gf %>%
  select(
    subj,
    gf_repeat_instance = redcap_repeat_instance,
    gf_date = gbl_func_date,
    gfs = gbl_func_soc_curr,
    gfr = gbl_func_role_curr
  )

# match GFS/GFR data to all behav data
dat_behav <- merge(dat_behav, dat_gf, by = "subj", all.x = TRUE)
```

## IQ
```{r, echo = FALSE, include = FALSE}
### TMB IQ  ------------
# Make sure to pull REDCap report "Test My Brain", to include tmb baseline date 
dat_iq <- read.csv("../data/nontask_data/other_data/step_iq_20250914.csv")
dat_iq <- clean_data(dat_iq)
dat_iq <- clean_rep_inst(dat_iq, "tmb_baseline_date")

# create dataframe with only target variables
dat_iq <- dat_iq %>%
  select(
    subj,
    iq_repeat_instance = redcap_repeat_instance, 
    iq_date = tmb_baseline_date, 
    iq = tmb_iq_estimate) 

# replace TMB values which are invalid (<70) with WTAR estimates
dat_iq[dat_iq$subj == "SP2014", "iq"] <- 98 #nonbinary, using Hispanic female estimate
dat_iq[dat_iq$subj == "SP2150", "iq"] <- 102
dat_iq[dat_iq$subj == "SP2199", "iq"] <- 99 #biracial, using Black estimate
dat_iq[dat_iq$subj == "SP2263", "iq"] <- 91

class(dat_iq$iq)

# match TMB-IQ data to all behav data
dat_behav <- merge(dat_behav, dat_iq, by = "subj", all.x = TRUE)
str(dat_behav)
```

## TOPX
```{r, echo = FALSE, include = FALSE}
### TOPX  ------------
# # read in aggregate TOPX data 
# dat_topx <- read.csv("../data/aggregated_data_20240802.csv")
# dat_topx_bl1 <- filter(dat_topx, dat_topx$visit == "BL1")
# dat_topx_bl1_wide <- dat_topx_bl1 %>%
#   pivot_wider(id_cols = c(subj, task, visit, date, 
#                           subj_type, d_context, d_expect),
#               names_from = trial_type,
#               values_from = c(n, numcorrect, rt_mean, rt_mdn, 
#                               accuracy, accuracy_adj, error),
#               names_vary = "slowest",
#               names_glue = "{trial_type}_{.value}")
# 
# # match TOPX data to all behav data
# dat_behav <- merge(dat_behav, dat_topx_bl1_wide, by = "subj", all.x = TRUE)

```

# Save behav dat file
```{r, echo = FALSE, include = FALSE}
# Save file ----------------
write.csv(dat_behav,
          file=paste0("/Users/arend103/Documents/umn_work/topx_analyses/topx_behav/data/step_behav_dat",
                      gsub("-", "", Sys.Date()), ".csv"), row.names = FALSE)

# save copy to DDM folder
# write.csv(dat_behav,file=paste0("/Users/arend103/Documents/umn_work/topx_analyses/topx_ddm/data/step_behav_dat_for_ddm_",gsub("-", "", Sys.Date()), ".csv"), row.names = FALSE)
```

# Descriptives - Pt/Ctl
```{r}
# number of patients and controls
table(dat_behav$group)

### create datasets by group ----------
dat_behav_pt <- subset(dat_behav, group == "Patient") 
dat_behav_ctl <- subset(dat_behav, group == "Control")
nPt <- nrow(dat_behav_pt)
nCtl <- nrow(dat_behav_ctl)
```

## Functions
### Count + % table
```{r, echo = FALSE}
# function for count, %, and NA table - 2 split (pt/ctl)
table_ct_pct_2split <- function(var) {
  
  # Calculate distribution and percentages for patients
  count_pt <- table(dat_behav_pt[[var]])
  pct_pt <- round((count_pt/nPt)*100, 0)
  na_pt <- sum(is.na(dat_behav_pt[[var]]))
  
  # Calculate distribution and percentages for controls
  count_ctl <- table(dat_behav_ctl[[var]])
  pct_ctl <- round((count_ctl/nCtl)*100, 0)
  na_ctl <- sum(is.na(dat_behav_ctl[[var]]))
  
  # Create a summary table
  summary <- tibble(
    Group = c(rep("Patient", length(count_pt) + 1), 
            rep("Control", length(count_ctl) + 1)),
    Variable = c(names(count_pt), "NA", 
             names(count_ctl), "NA"),
    Count = c(as.numeric(count_pt), na_pt,
            as.numeric(count_ctl), na_ctl),
    Percentage = c(as.numeric(pct_pt), NA,
                 as.numeric(pct_ctl), NA)
    )
  summary
}
```

### M + SD table
```{r, echo = FALSE}
# function for mean, sd, and NA table - 2 split (pt/ctl)
table_m_sd_2split <- function(var) {
  
  # Calculate mean and standard deviation for patients
  mean_pt <- mean(dat_behav_pt[[var]], na.rm = TRUE)
  sd_pt <- sd(dat_behav_pt[[var]], na.rm = TRUE)
  
  # Calculate mean and standard deviation for controls
  mean_ctl <- mean(dat_behav_ctl[[var]], na.rm = TRUE)
  sd_ctl <- sd(dat_behav_ctl[[var]], na.rm = TRUE)
  
  # Count missing values for patients and controls
  na_pt <- sum(is.na(dat_behav_pt[[var]]))
  na_ctl <- sum(is.na(dat_behav_ctl[[var]]))
  
  # Create a summary table
  summary <- tibble(
    Group = c("Patient", "Control"),
    Mean = c(round(mean_pt, 2), round(mean_ctl, 2)),
    SD = c(round(sd_pt, 2), round(sd_ctl, 2)),
    Missing_Values = c(na_pt, na_ctl)  # Add missing values count
  )
  summary %>% mutate(across(where(is.numeric), as.character)) # this is needed to report two digits
}
```

### BPRS table - reg factors
```{r, echo = FALSE}
table_bprs <- function(data) {
  summary <- tibble(
    Variable = c("BPRS Positive", "BPRS Negative", 
                 "BPRS Disorganized", "BPRS Mania", "BPRS Depression"),
    Mean = c(
      round(mean(data$bprs_positive, na.rm = TRUE), 2),
      round(mean(data$bprs_negative, na.rm = TRUE), 2),
      round(mean(data$bprs_disorganized, na.rm = TRUE), 2),
      round(mean(data$bprs_mania, na.rm = TRUE), 2),
      round(mean(data$bprs_depression, na.rm = TRUE), 2)),
    SD = c(
      round(sd(data$bprs_positive, na.rm = TRUE), 2),
      round(sd(data$bprs_negative, na.rm = TRUE), 2),
      round(sd(data$bprs_disorganized, na.rm = TRUE), 2),
      round(sd(data$bprs_mania, na.rm = TRUE), 2),
      round(sd(data$bprs_depression, na.rm = TRUE), 2))
    )
  summary
}
```

### BPRS - W/S average
```{r, echo = FALSE}
table_bprs_ws <- function(data) {
  summary <- tibble(
    Variable = c("BPRS Positive", "BPRS Negative", 
                 "BPRS Disorganized", "BPRS Mania", "BPRS Depression"),
    Mean = c(
      round((mean(data$bprs_pos_ws, na.rm = TRUE)/5), 2),
      round((mean(data$bprs_neg_ws, na.rm = TRUE)/3), 2),
      round((mean(data$bprs_disorg_ws, na.rm = TRUE)/6), 2),
      round((mean(data$bprs_mania_ws, na.rm = TRUE)/5), 2),
      round((mean(data$bprs_dep_anx_ws, na.rm = TRUE)/4), 2)),
    SD = c(
      round((sd(data$bprs_pos_ws, na.rm = TRUE)/5), 2),
      round((sd(data$bprs_neg_ws, na.rm = TRUE)/3), 2),
      round((sd(data$bprs_disorg_ws, na.rm = TRUE)/6), 2),
      round((sd(data$bprs_mania_ws, na.rm = TRUE)/5), 2),
      round((sd(data$bprs_dep_anx_ws, na.rm = TRUE)/4), 2))
    )
  summary
}
```

### BPRS - W/S total
```{r, echo = FALSE}
table_bprs_ws_tot <- function(data) {
  summary <- tibble(
    Variable = c("BPRS Positive", "BPRS Negative", 
                 "BPRS Disorganized", "BPRS Mania", "BPRS Depression"),
    Mean = c(
      round(mean(data$bprs_pos_ws, na.rm = TRUE), 2),
      round(mean(data$bprs_neg_ws, na.rm = TRUE), 2),
      round(mean(data$bprs_disorg_ws, na.rm = TRUE), 2),
      round(mean(data$bprs_mania_ws, na.rm = TRUE), 2),
      round(mean(data$bprs_dep_anx_ws, na.rm = TRUE), 2)),
    SD = c(
      round(sd(data$bprs_pos_ws, na.rm = TRUE), 2),
      round(sd(data$bprs_neg_ws, na.rm = TRUE), 2),
      round(sd(data$bprs_disorg_ws, na.rm = TRUE), 2),
      round(sd(data$bprs_mania_ws, na.rm = TRUE), 2),
      round(sd(data$bprs_dep_anx_ws, na.rm = TRUE), 2))
    )
  summary
}
```

### SANS-SAPS avg
```{r}
table_sanssaps <- function(data) {
  summary <- tibble(
    Variable = c("SANS-SAPS Total", "SAPS Positive", "SANS Negative", 
                 "SANS Negative v2",  "SANS-SAPS Disorganized"),
    Mean = c(
      round(mean(data$sanssaps_total, na.rm = TRUE), 2),
      round(mean(data$sanssaps_pos_tib2, na.rm = TRUE), 2),
      round(mean(data$sanssaps_neg_tib2, na.rm = TRUE), 2),
      round(mean(data$sanssaps_neg_tib2b, na.rm = TRUE), 2),
      round(mean(data$sanssaps_disorg_tib2, na.rm = TRUE), 2)),
    SD = c(
      round(sd(data$sanssaps_total, na.rm = TRUE), 2),
      round(sd(data$sanssaps_pos_tib2, na.rm = TRUE), 2),
      round(sd(data$sanssaps_neg_tib2, na.rm = TRUE), 2),
      round(sd(data$sanssaps_neg_tib2b, na.rm = TRUE), 2),
      round(sd(data$sanssaps_disorg_tib2, na.rm = TRUE), 2))
    )
  summary
}
```

## Demographics
### Gender
```{r}
#gender: 1=man, 2=woman, 3=nonbinary/other
table_ct_pct_2split("gender")

# 1 = cis man, 2 = cis woman, 3 = trans/enby
table_ct_pct_2split("gender_3split")

# 1=cis man, 2=cis woman, 3=trans man, 4=trans woman, 5 = enby
table_ct_pct_2split("gender_5split")

# lgbtq
table_ct_pct_2split("lgbtq")

# pt percentage NA (6/73)*100
```


### Age
```{r}
table_m_sd_2split("age")
```

### Race
```{r}
# race: 0=white, 1=black/african am, 2=indigenous am/alaska native, 3=asian/asian am, 4=native hawaiian/pacific islander, 5=multiracial, 6=unknown/other
table_ct_pct_2split("race")
```

### Race 5 split
```{r}
# race: 0=white, 1=black/african am, 2=indigenous am/alaska native, 3=asian/asian am, 4=native hawaiian/pacific islander, 5=multiracial/other
table_ct_pct_2split("race_5split")
```

### Hispanicity
```{r}
# manually changed SP1030 = yes, SP1097 = no, SP1099 = no
table_ct_pct_2split("hispanic")
```

### Personal Education (Highest)
```{r}
table_m_sd_2split("subj_education")
```

### Caregiver Education (Highest)
```{r}
table_m_sd_2split("hi_caregiver_education")
```

### IQ
```{r}
table_m_sd_2split("iq")
```

## Clinical 
### Diagnosis - 7 split
```{r}
# dx: 0=control, 1=Sz, 2=SzAff, 3=Szform, 4=NOS, 5=BP, 6=MDD, 7=other
table_ct_pct_2split("primary_dx")
```

### Diagnosis - 5 split
```{r}
# dx: 0 = control, 1 = Sz/SzAff/Szform, 2 = BP, 3 = MDD, 4 = NOS/other
table_ct_pct_2split("dx_5split")
```

### BPRS total
```{r}
table_m_sd_2split("bprs_total")
```

### BPRS - Pt Wilson-Sponheim Factor Average
```{r}
#table_bprs(dat_behav_pt)
table_bprs_ws(dat_behav_pt)
```

### BPRS - Pt Wilson-Sponheim Factor Total
```{r}
#table_bprs(dat_behav_pt)
table_bprs_ws_tot(dat_behav_pt)
```
### BPRS - Ctl Wilson-Sponheim Factor Average
```{r}
#table_bprs(dat_behav_ctl)
table_bprs_ws(dat_behav_ctl)
```

### BPRS - Ctl Wilson-Sponheim Factor Total
```{r}
#table_bprs(dat_behav_ctl)
table_bprs_ws_tot(dat_behav_ctl)
```

### SANS-SAPS - Pt
```{r}
table_sanssaps(dat_behav_pt)
# in paper, report negative, not negative v2 (v2 eliminates shared variables)
```


### Duration of untreated psychosis
```{r}
table_m_sd_2split("duration_of_psychosis")
```

## Group differences
```{r}
### demo group differences --------------
# gender - report fisher
chisq.test(dat_behav$gender, dat_behav$group)
fisher.test(dat_behav$gender, dat_behav$group)
chisq.test(dat_behav$gender_3split, dat_behav$group)
fisher.test(dat_behav$gender_3split, dat_behav$group)



# age
# welch's two-sample t-test - report this
t.test(age ~ status_ppt_type, data = dat_behav)
# student t-test (assumes equal variances)
t.test(age ~ status_ppt_type, data = dat_behav, var.equal = TRUE)
# test group variances to see if t-test are appropriate
leveneTest(dat_behav$age, dat_behav$status_ppt_type, data = dat_behav)
var.test(age ~ status_ppt_type, data = dat_behav)
# wilcox aka mann-whitney U test
wilcox.test(age ~ status_ppt_type, data = dat_behav)

# race - report fisher
chisq.test(dat_behav$race, dat_behav$group)
fisher.test(dat_behav$race, dat_behav$group)

# ethnicity
chisq.test(dat_behav$hispanic, dat_behav$group)

# participant education
t.test(subj_education ~ status_ppt_type, data = dat_behav) #welch's
leveneTest(dat_behav$subj_education, dat_behav$status_ppt_type, data = dat_behav)
var.test(subj_education ~ status_ppt_type, data = dat_behav)
wilcox.test(subj_education ~ status_ppt_type, data = dat_behav)

# caregiver education
t.test(hi_caregiver_education ~ status_ppt_type, data = dat_behav) # welch's

# IQ
leveneTest(dat_behav$iq, dat_behav$status_ppt_type, data = dat_behav)
shapiro.test(dat_behav$iq)

t.test(iq ~ status_ppt_type, data = dat_behav) #welch's
wilcox.test(iq ~ status_ppt_type, data = dat_behav)

# BPRS - total
leveneTest(dat_behav$bprs_total, dat_behav$status_ppt_type, data = dat_behav)
t.test(bprs_total ~ status_ppt_type, data = dat_behav)
wilcox.test(bprs_total ~ status_ppt_type, data = dat_behav)

# BPRS - positive
leveneTest(dat_behav$bprs_pos_ws, dat_behav$status_ppt_type, data = dat_behav)
t.test(bprs_pos_ws ~ status_ppt_type, data = dat_behav)
wilcox.test(bprs_pos_ws ~ status_ppt_type, data = dat_behav)

# BPRS - negative
leveneTest(dat_behav$bprs_neg_ws, dat_behav$status_ppt_type, data = dat_behav)
t.test(bprs_neg_ws ~ status_ppt_type, data = dat_behav)
wilcox.test(bprs_neg_ws ~ status_ppt_type, data = dat_behav)

# BPRS - disorg
leveneTest(dat_behav$bprs_disorg_ws, dat_behav$status_ppt_type, data = dat_behav)
t.test(bprs_disorg_ws ~ status_ppt_type, data = dat_behav)
wilcox.test(bprs_disorg_ws ~ status_ppt_type, data = dat_behav) 

# BPRS - mania
leveneTest(dat_behav$bprs_mania_ws, dat_behav$status_ppt_type, data = dat_behav)
t.test(bprs_mania_ws ~ status_ppt_type, data = dat_behav)
wilcox.test(bprs_mania_ws ~ status_ppt_type, data = dat_behav) 

# BPRS - depr/anx
leveneTest(dat_behav$bprs_dep_anx_ws, dat_behav$status_ppt_type, data = dat_behav)
t.test(bprs_dep_anx_ws ~ status_ppt_type, data = dat_behav)
wilcox.test(bprs_dep_anx_ws ~ status_ppt_type, data = dat_behav) 
```


# Descriptives - Aff/NonAff
```{r}
# number of controls, Non-Affective, Affective
# 0 = Control, 1 = Non-Affective, 2 = Affective
table(dat_behav$dx_affect)

### create datasets by group ----------
# run above: dat_behav_ctl <- subset(dat_behav, group == "Control")
# run above: nCtl <- nrow(dat_behav_ctl)
dat_behav_na <- subset(dat_behav, dx_affect == 1) 
dat_behav_aff <- subset(dat_behav, dx_affect == 2) 
nPt_na <- nrow(dat_behav_na)
nPt_aff <- nrow(dat_behav_aff)
```

## Functions
### Count + % table
```{r, echo = FALSE}
# function for count, %, and NA table - 3 split (NA/Aff/ctl)
table_ct_pct_3split <- function(var) {
  
  # Calculate distribution and percentages for NA patients
  count_na <- table(dat_behav_na[[var]])
  pct_na <- round((count_na/nPt_na)*100, 2)
  na_na <- sum(is.na(dat_behav_na[[var]]))
  
  # Calculate distribution and percentages for Aff patients
  count_aff <- table(dat_behav_aff[[var]])
  pct_aff <- round((count_aff/nPt_aff)*100, 2)
  na_aff <- sum(is.na(dat_behav_aff[[var]]))
  
  # Calculate distribution and percentages for controls
  count_ctl <- table(dat_behav_ctl[[var]])
  pct_ctl <- round((count_ctl/nCtl)*100, 2)
  na_ctl <- sum(is.na(dat_behav_ctl[[var]]))
  
  # Create a summary table
  summary <- tibble(
    Group = c(rep("NA", length(count_na) + 1), 
              rep("Aff", length(count_aff) + 1), 
              rep("Control", length(count_ctl) + 1)),
    Variable = c(names(count_na), "NA", 
                 names(count_aff), "NA",
                 names(count_ctl), "NA"),
    Count = c(as.numeric(count_na), na_na,
              as.numeric(count_aff), na_aff,
              as.numeric(count_ctl), na_ctl),
    Percentage = c(as.numeric(pct_na), NA,
                   as.numeric(pct_aff), NA,
                   as.numeric(pct_ctl), NA)
    )
  summary
}
```

### M + SD table
```{r, echo = FALSE}
# function for mean, sd, and NA table - 3 split (NA/Aff/ctl)
table_m_sd_3split <- function(var) {
  
  # Calculate mean and standard deviation for non-affective
  mean_na <- mean(dat_behav_na[[var]], na.rm = TRUE)
  sd_na <- sd(dat_behav_na[[var]], na.rm = TRUE)
  
  # Calculate mean and standard deviation for affective
  mean_aff <- mean(dat_behav_aff[[var]], na.rm = TRUE)
  sd_aff <- sd(dat_behav_aff[[var]], na.rm = TRUE)
  
  # Calculate mean and standard deviation for controls
  mean_ctl <- mean(dat_behav_ctl[[var]], na.rm = TRUE)
  sd_ctl <- sd(dat_behav_ctl[[var]], na.rm = TRUE)
  
  # Count missing values
  na_na <- sum(is.na(dat_behav_na[[var]]))
  na_aff <- sum(is.na(dat_behav_aff[[var]]))
  na_ctl <- sum(is.na(dat_behav_ctl[[var]]))
  
  # Create a summary table
  summary <- tibble(
    Group = c("Non-Affective", "Affective", "Control"),
    Mean = c(round(mean_na, 2), 
                 round(mean_aff, 2),
                 round(mean_ctl, 2)),
    SD = c(round(sd_na, 2), 
               round(sd_aff,2),
               round(sd_ctl, 2)),
    Missing_Values = c(na_na, na_aff, na_ctl)
    )
  summary
}
```

## Demographics
### Gender
```{r}
#gender: 1=man, 2=woman, 3=nonbinary/other
table_ct_pct_3split("gender")
```

### Age
```{r}
table_m_sd_3split("age")
```

### Race
```{r}
# race: 0=white, 1=black/african am, 2=indigenous am/alaska native, 3=asian/asian am, 4=native hawaiian/pacific islander, 5=multiracial, 6=unknown/other
table_ct_pct_3split("race")
```

### Hispanicity
```{r}
table_ct_pct_3split("hispanic")
```

### Caregiver Education (Highest)
```{r}
table_m_sd_3split("hi_caregiver_education")
```

## Clinical
### Diagnosis - 3 split (Affective/Non-affective)
```{r}
table_ct_pct_3split("dx_affect")
```

### Duration of untreated psychosis
```{r}
table_m_sd_3split("duration_of_psychosis")
```

### BPRS - NA
```{r}
table_bprs(dat_behav_na)
```

### BPRS - Aff
```{r}
table_bprs(dat_behav_aff)
```

### BPRS - Ctl
```{r}
table_bprs(dat_behav_ctl)
```
