---
title: "TOPX Learning Effects"
author: "Jessica Arend"
date: "2024-10-02"
output: html_document
---

# Between BL1-BL2
# Load in and prep data/workspace
```{r, echo = FALSE}
# clear workspace
rm(list=ls())

# set working directories
pacman::p_load(dplyr, stringr, ggplot2, tidyr)

# load in data
dat <- read.csv("~/Documents/UMN Work & General/TOPX analyses/mri aggregated data/aggregated_data_20240802.csv")
View(dat)

# subset by visit
dat_bl1 <- dat %>% 
  filter(visit == "BL1") 
dat_bl2 <- dat %>% 
  filter(visit == "BL2") 
dat_6m <- dat %>% 
  filter(visit == "6M") 
dat_7m <- dat %>% 
  filter(visit == "7M") 
dat_12m <- dat %>% 
  filter(visit == "12M") 

# created new df with merged BL1 and BL2
dat_bl1_rev <- subset(dat_bl1, (subj %in% dat_bl2$subj))
dat_bl2_rev <- subset(dat_bl2, (subj %in% dat_bl1$subj))
dat_bls <- rbind(dat_bl1_rev, dat_bl2_rev)

# recode visit with numerical values
# dat_bls <- dat_bls %>% 
#   mutate(visit_num =  case_when(visit == "BL1" ~ 1,
#                        visit == "BL2" ~ 2))
```

# Learning effects of error rates between BL1-BL2
## Plot error rates
```{r, echo = FALSE}
ggplot(dat_bls, aes(x= trial_type, y=error, group = interaction(subj_type,visit))) + 
  stat_summary(fun="mean", geom="line", aes(color = subj_type, linetype = visit), size = 1.5) + stat_summary(fun="mean", geom="point", aes(color = subj_type), size = 3) + stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = subj_type), width = 0.1, size = 1) + 
  labs(title = "Error Rates", 
       y = "Error Rate", 
       x = "Trial Type", 
       color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(0.05, .35)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Control", "EP")) + 
  scale_linetype_manual(values = c("solid", "dotted"), 
                        labels = c("Visit 1", "Visit 2")) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
```

# T-tests of error by trial, all participants
```{r}
# create function to subset by trial and run t-test across all ppts
t_err_trial <- function(dat, trial) {
  dat_subset <- dat[dat$trial_type == trial, ]
  result <- t.test(dat_subset$error ~ dat_subset$visit)
  return(result)}

# run t-tests with function
t_err_trial(dat_bls, "AX")
t_err_trial(dat_bls, "AY")
t_err_trial(dat_bls, "BX")
t_err_trial(dat_bls, "BY")
```

# T-tests of error by trial and group
```{r}
# create function to subset by group and trial and run t-test
t_err_trial_grp <- function(dat, trial, group) {
  dat_subset <- dat[dat$trial_type == trial & dat$subj_type == group, ]
  result <- t.test(dat_subset$error ~ dat_subset$visit)
  return(result)}

t_err_trial_grp(dat_bls, "AX", "Control")
t_err_trial_grp(dat_bls, "AX", "Patient")
t_err_trial_grp(dat_bls, "AY", "Control")
t_err_trial_grp(dat_bls, "AY", "Patient")
t_err_trial_grp(dat_bls, "BX", "Control")
t_err_trial_grp(dat_bls, "BX", "Patient")
t_err_trial_grp(dat_bls, "BY", "Control")
t_err_trial_grp(dat_bls, "BY", "Patient")

```

# Learning effects of reaction times between BL1-BL2
## Plot reaction times
```{r, echo = FALSE}
ggplot(dat_bls, aes(x= trial_type, y=rt_mdn, group = interaction(subj_type,visit))) + stat_summary(fun="mean", geom="line", 
aes(color = subj_type, linetype = visit), size = 1.5) + stat_summary(fun="mean", geom="point", aes(color = subj_type), size = 3) + stat_summary(fun.data = mean_se, geom = "errorbar", aes(color = subj_type), width = 0.1, size = 1) + 
  labs(title = "Reaction Times (RT)", 
       y = "RT (ms)", 
       x = "Trial Type", color = "Group", linetype = "Visit") +
  coord_cartesian(ylim = c(280, 600)) +
  scale_color_manual(values=c("#2C737F", "#AA8829"),
                     labels=c("Control", "EP")) + 
  scale_linetype_manual(values = c("solid", "dotted"), 
                        labels = c("Visit 1", "Visit 2")) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black"), 
        axis.text = element_text(face="bold", size = 16),
        axis.title = element_text(face="bold", size = 20),
        legend.text = element_text(face="bold", size = 14),
        legend.title = element_text(face = "bold", size = 16),
        plot.title = element_text(face = "bold", size = 22))
```

# T-tests of error by trial, all participants
```{r}
# create function to subset by trial and run t-test across all ppts
t_rt_trial <- function(dat, trial) {
  dat_subset <- dat[dat$trial_type == trial, ]
  result <- t.test(dat_subset$rt_mdn ~ dat_subset$visit)
  return(result)}

# run t-tests with function
t_rt_trial(dat_bls, "AX")
t_rt_trial(dat_bls, "AY")
t_rt_trial(dat_bls, "BX")
t_rt_trial(dat_bls, "BY")
```

# T-tests of error by trial and group
```{r}
# create function to subset by group and trial and run t-test
t_rt_trial_grp <- function(dat, trial, group) {
  dat_subset <- dat[dat$trial_type == trial & dat$subj_type == group, ]
  result <- t.test(dat_subset$rt_mdn ~ dat_subset$visit)
  return(result)}

t_rt_trial_grp(dat_bls, "AX", "Control")
t_rt_trial_grp(dat_bls, "AX", "Patient")
t_rt_trial_grp(dat_bls, "AY", "Control")
t_rt_trial_grp(dat_bls, "AY", "Patient")
t_rt_trial_grp(dat_bls, "BX", "Control")
t_rt_trial_grp(dat_bls, "BX", "Patient")
t_rt_trial_grp(dat_bls, "BY", "Control")
t_rt_trial_grp(dat_bls, "BY", "Patient")

```

# Between blocks
# Load in and prep data/workspace
```{r, echo = FALSE}
# load in data
dat_blocks <- read.csv("~/Documents/UMN Work & General/TOPX analyses/mri aggregated data/aggregated_data_byblock_20241002.csv")
View(dat_blocks)

# subset by visit
dat_bl1_blocks <- dat_blocks %>% 
  filter(visit == "BL1") 
dat_bl2_blocks <- dat_blocks %>% 
  filter(visit == "BL2") 
```